---
title: "Module 1 - Exploratory Data Analysis"
output: 
  learnr::tutorial:
    css: css/style.css
runtime: shiny_prerendered
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(DT)
library(plotly)
library(shiny)
library(waffle)
library(treemap)
library(gt)
library(janitor)
library(yaml) 
library(Hmisc)
knitr::opts_chunk$set(echo = FALSE,message = FALSE,warning=FALSE)

choices<-c("Point"="point",
"Jittered Plot"="jitter",
"Boxplot"="boxplot",
"Violin Plot"="violin",
"Histogram"="histogram",
"Frequency Polygon"="freq_poly",
"Density Plot"="density",
"Caterpillar Plot"="caterpillar",
"Bar Chart"="bar",
"Dynamite Plot"="dynamite",
"Mean + Error bar"="errorbar",
"Cumulative Density Plot"="cdf",
"Normal QQ Plot"="qq")

choices2<-c("Bar Chart"="bar",
"Stacked Bar Chart"="stacked_bar",
"Pie Chart"="pie",
"Dotplot"="count",
"Waffle Plot"="waffle",
"Tree Map"="tree_map")

plot_theme<-theme_light()+
  theme(plot.title=element_text(face =2,size=18),
        axis.title=element_text(face =2,size=14),
        plot.subtitle=element_text(face =3,size=14),
        axis.text=element_text(face =1,size=12),
        strip.text=element_text(face =1,size=12))


```

```{r,context="data",include=FALSE}

ghg_2022<-readRDS("data/ghg_summary_data.RDS") %>% droplevels()
ghg_snap<-readRDS("data/ghg_snap.RDS")
ghg_summary<-readRDS("data/ghg_summary.RDS")
ghg_allyears<-readRDS("data/ghg_summary_data_all.RDS")
ghg_una<-readRDS("data/ghg_una.RDS")
ts<-readRDS("data/ghg_ts.RDS")
ghg_missing<-readRDS("data/ghg_missing.RDS")

ghg_2022 %>%
  ungroup() %>%
      summarise("Number of Counties"=n(),
                Mean=mean(Annual.CO2.Total)%>%round(1),
                Median=median(Annual.CO2.Total)%>%round(1),
                "Trimmed mean"=mean(Annual.CO2.Total,trim = 0.05)%>%round(1),
                 "Geometric mean"=psych::geometric.mean(Annual.CO2.Total)%>%round(1),
                "Harmonic mean"=psych::harmonic.mean(Annual.CO2.Total)%>%round(1),
                 Sum=sum(Annual.CO2.Total)%>%round(),
                "Minimum"=min(Annual.CO2.Total)%>%round(1),
                "Maximum"=max(Annual.CO2.Total)%>%round(1),
                "Range"=(max(Annual.CO2.Total)-min(Annual.CO2.Total))%>%round(1),
                "Standard Deviation"=sd(Annual.CO2.Total)%>%round(1),
                "Variance"=var(Annual.CO2.Total)%>%round(),
                "Coefficient of Variation"=scales::percent(sd(Annual.CO2.Total)/mean(Annual.CO2.Total)),
                "Lower Quartile"=quantile(Annual.CO2.Total,0.25)%>%round(1),
                "Upper Quartile"=quantile(Annual.CO2.Total,0.75)%>%round(1),
                "Inter Quartile Range"=diff(quantile(Annual.CO2.Total,c(0.25,0.75)))%>%round(1),
                "Median Absolute Deviation"=mad(Annual.CO2.Total)%>%round(1),
                "Kurtosis"=agricolae::kurtosis(Annual.CO2.Total)%>%round(2),
                "Skewness"=agricolae::skewness(Annual.CO2.Total)%>%round(2),
                "% of Values Over Threshold (100kt)"=scales::percent(mean(Annual.CO2.Total>1000)),
                )%>%
  gather(Statistic,Value) ->table_stat

```

## Introduction

We are going to start this course with a module covering what may be the
most important, but most commonly overlooked, step in statistical
analysis.

Exploratory data analysis is the process of exploring, visualising and
summarising our data before we move to any formal statistical analysis.
While it may be tempting to skip to the end and start immediately
fitting models or testing hypotheses to get the "results", going into
these blind may lead to completely incorrect conclusions, overlooking
important findings, or missing key caveats that would be needed. The aim
of this process should be to help us to understand the data we are
working with, uncover general patterns and trends in our data, identify
possible data errors & inform choices for our future modelling about the
shape of distributions. The most important part of this phase of
analysis is to truly consider it as an exploration, rather than a box
ticking exercises. There are no wrong approaches, but you may learn
different things from being willing and able to explore your data in
different ways. The goal is simply to learn as much as possible about
your data and starting to establish the links between how your data
links to reality, and how your data can mapped against your research
questions.

This session we will cover:

-   Data familiarisation and checking

-   Exploratory methods for univariate analysis to identify possible
    errors or outliers, and to understand the shape, central tendency
    and distribution of the variables

-   Exploratory Methods to identify relationships between two or more
    variables

-   Data manipulations & transformations

Most of the methods we cover in this session are likely to be somewhat
familiar to you. Rather than covering these methods in detail, instead
we will try to summarise the strengths & weaknesses of these methods and
under which circumstances they would be most useful.

In the video I will walk through the process of an exploration of the
dataset that we will use throughout this workbook - following the trail
of starting out with my individual variables, and then linking them
together - and making sure I account for the important points I am
noticing about the variables, and their relationship as I talk through.

![](https://youtu.be/BL3WdUWm1Q8)

A good data explorer needs to be able to comfortably work with your
data - manipulating it where needed by summarising to different levels,
correct mistakes, or derive new variables - and to be able to produce a
flexible range of different visualisations. You might be able to do some
of this using Excel or Google Sheets, but to do it well and to do it in
a reproducible and flexible way you will likely need to be using some
other software tools like R or Python. In the video, and in the
exercises, I used a simple interactive data exploration tool - built
using R - to help me work through quickly without needing to do any live
coding; but at the expense of having full flexibility over everything I
might want to do.

In these workbooks all output produced will come from the use of R, and
all of the code used in the workbooks can be found on GitHub here:
<https://github.com/stats4sd/Stats_ReIntro_M1>. But the principles are
identical regardless of which data analysis tool is being used.Whichever
tool is being used for data analysis - it will require that the user is
familiar enough with it to be able to generate any sort of summary or
plot or transformation that they might suddenly think could be useful!

## Data being used in this workbook

In this session we are going to explore a public dataset produced by the
UK government summarising greenhouse gas emissions over time from 2005
to 2022, at local authority ('county') level, attributed to various
different sources. This data is updated on an annual basis - you can
read more about it here including reading through the metadata in full,
or you could download a more recent version containing data from 2023
and beyond should it be available when you are reading this!

<https://www.gov.uk/government/statistics/uk-local-authority-and-regional-greenhouse-gas-emissions-statistics-2005-to-2022>

<https://assets.publishing.service.gov.uk/media/667ad86497ea0c79abfe4bfd/2005-2022-local-authority-ghg-emissions-csv-dataset.csv>

## Understanding & Checking the Data

... it may sound obvious but the first thing that needs to be done when
working with a dataset is to look at it! This is particularly the case
if as part of your research you have inherited your data from a
secondary source, or it has come from an automated process, rather than
being data that you have collected yourself and are likely to be more
familiar with already. And even if you have collected the data yourself
there is still a pretty high probability that you have made a mistake
somewhere along the way, in the data entry or compilation phases.

When first looking through our data there are a lot of different things
we want to be considering:

-   What does each column contain?

-   Are the column names coherent for us to understand the data?

-   Has our data been interpreted appropriately by our software - are
    there perhaps issues where metadata has been confused for data, or
    columns contain multiple variables?

-   What units are each of the columns recorded in?

-   Are the numbers consistent with those units?

-   Do the values in that column make some sense when tied to your own
    understanding of reality? Are there obviously incorrect high or low
    values; or zeros when there should be values?

-   Are there missing values? And if so can we understand why there are
    missing values? Is the data missing because data was not collected,
    or not relevant, or corrupted in some way?

-   Have codes been used to denote different levels of a categorical
    variables? Can you de-code these codes? If not - do we have the
    relevant metadata that can help us interpret them?

-   What does each row contain?

-   Does the number of rows match what we expect (e.g. if we are
    expecting one row of data per hour, do we always have 24
    observations per day?)

-   Do we have any duplicated or obsolete rows or columns in the data?

While a lot of this may seem fairly basic, there are countless examples
of data analysis mistakes, some minor some extremely major, made because
analysts simply did not spend the time to look at and understand the
data they were working with and ploughed straight through into the
analysis. e.g. <https://www.bbc.co.uk/news/technology-54423988> ;
<https://stats4sd.org/blog/22> ;
<http://svmiller.com/blog/2020/04/reinhart-rogoff-ten-years-later-replication/>
).

If the data collection is your responsibility - then part of the process
of data collection and compilation should be data documentation and
validation - ensuring that any mistakes in data entry are identified
robustly and that metadata is produced allowing others to be able to
inherit your data in the future and understand what each variable
corresponds to, how it was collected and what it means.

### Looking at the UK Greenhouse Gas Data

If you downloaded the full dataset from the previous link above you
would see that it has 533,016 rows. The first step, let's just take a
look at the data - the first 1000 rows are shown below:

```{r}
DT::datatable(ghg_snap)
```

From a quick inspection visually of the first few hundred rows of this
data we can actually learn quite a lot and start to answer some of the
questions we may have.

The first thing you might notice is that there seems to be a lot of data
from Hartlepool, and not a lot of data from anywhere else! You might be
worried that data has been incorrectly duplicated.

Closer inspection of the data shows why the data looks like this - we
can see that data is stored as one row per gas per sub-sector per year
per local authority. And we have three different gases (CO2,CH4 and
N20), 32 different sub sectors, and 18 years of data. So that's
potentially up to 1,728 rows per local authority.

So the first 1000 rows will indeed all come from Hartlepool, since the
Local Authority code "E06000001" happens to be the first one
alphabetically. Having data with this structure also tells us that we
will need to be comfortable with manipulating the data between different
levels - if we wanted to look at the total emissions per year, or within
a sector, or correlate different sources/gases against each other then
we would need to be able to manipulate our dataset in various different
ways to allow us to do this.

The column names and units in the data are relatively transparent as to
what they mean. The acronyms used are all relatively standard - some of
the less familiar ones may be "LA"="Local Authority"; "GHG"="Greenhouse
Gas"; "kt"="Kilo-tonnes";"CO2e"="CO2 equivalent". And, the UK government
contains extensive metadata and documentation that would be able to help
us if we got stuck here.

Something that is very clear is that manually checking through 533,016
rows of data sounds like a terrible idea!

Within R, and within most equivalent software, there will be commands to
explore the structure, or summary of the data, and this can help us to
identify other potentially important points about the data as a quick
initial snapshot.

```{r}
ghg_summary
```

There are a lot of numbers in there - but there is actually a lot of
really useful insights we can make, that could help us to understand the
data:

-   There are 414 rows with missing values for **population** and
    **area** of the county.
-   There are also 414 rows in the data where **country** is
    "Unallocated" to any of the countries of the UK. It would seem
    logical that these may be the same rows - but let us check to make
    sure!

```{r}
ghg_missing %>% gt()
```

-   If we look more at data restricted to show only the unallocated rows
    the "local authority column" we can see that these are definitely
    not local authorities! In fact this column instead they provide a
    very short description of the context - "Large elec users" or
    "Unallocated consumption". Nearly all the rows for these unallocated
    entries are for the Industry or Domestic sectors although there are
    some which relate to Waste or LULUCF
    (<https://unfccc.int/topics/land-use/workstreams/land-use--land-use-change-and-forestry-lulucf>).
    The existence of these rows, attributing emissions to unknown
    sources, will mean we have to think carefully about what to do with
    them when producing summaries of our data as there may be a number
    of different approaches we could consider.

```{r}
DT::datatable(ghg_una)
```

-   There are negative values for the column
    "Territorial.emissions..kt.CO2e". We may not have been expecting
    that! In these cases we will have estimates linked to carbon
    offsetting schemes. So this column is dealing with *net* emissions,
    and this is an important point for us to understand. And it is good
    that we have identified these before we did something which would
    have been impossible with negative data values, like taking a log
    transformation for example.

-   There are **not** the same number of rows for each of the
    sub-sectors. Although some of this could be explained by the
    unallocated sources the disparity definitely accounts for a lot more
    than 414 rows! So we might then start to explore whether we have
    data for each sub sector within each year, for each of the three
    greenhouse gases and for each country.

**Extension Exercise** Download the raw data from the link in the
previous section if you have not done so already, and see if you can
explore the data to understand why some sub-sectors have more rows than
other.

### Data Aggregation & Restriction

To give ourself a slightly smaller task to take on in this workbook, we
will focus in on aggregated subset of the data designed to meet the
needs of a specific research question.

For the remainder of this module we will focus on the relationships that
exist between the CO2 emissions and the geographic characteristics of
each local authority from the most recently available data.

This will require three (or four) separate steps: - Subset the data so
that it only includes data from the most recent year - 2022 - Subset the
data so that it only contains data from the greenhouse gas of interest -
CO2 - Aggregate the data so that for each local authority we calculate
the sum of emissions across all of the sources accounted for - Being
careful with the population and area variables in our aggregation
process! We want to retain this variables in our new data set, but we do
not want to sum these across all of the different sources, as the values
of area and population are fixed for each local authority within each
year

Below you can explore the aggregated of the data showing the total CO2
emissions for each county from 2022 (`Annual.CO2.Total`), which is the
one I used throughout the video for this module.

```{r}
DT::datatable(ghg_2022)
```

## Exploring One Numeric Variable

### Summarising

Summary statistics help us to get a better idea of the variability, the
central tendency, the extreme values, and the distribution and relative
frequency of the values. Let's start with the CO2 emissions variable,
which is a continuous numeric variable and consider as many different
ways we can think of for producing summaries:

```{r sum_table}

  gt::gt(table_stat) %>%
  gt::fmt_number(columns = 2)
  
```

From the table you are hopefully fairly familiar with most of the
quantities mentioned; although I expect there will be some you will have
heard of but perhaps don't really understand, and maybe some with which
you are less familiar with.

The interactive table below provides a summary of how you might intepret
each of these numbers, and what the statistics can be used for. In
general these statistics will fall into one of 4 categories -

-   **frequency** - these statistics will tell us how much data is
    available, and how frequent specific events may be

-   **central tendency** - often referred to as "average", these
    statistics will tell us something about the 'middle' of our data, or
    the most likely values to occur in the data.

-   **variability** - these statistics will quantify how much deviation
    there is in our values away from the 'middle' of the data, and help
    us to the variation that exists within the data

-   **distribution** - these statistics will help us to understand the
    shape of the data. Are values equally spread above or below the
    average? Are there outliers and how to way degree do these fall
    outside of expected ranges? Is there a single peak around the
    "average" or are there multiple peaks?

```{r method_ui, context="render",echo=FALSE}

  selectInput("method",label = "Select Summary Method",
              choices =table_stat$Statistic)
 
  gt::gt_output("table2")

```

```{r method_server,context="server"}

output$table2<-render_gt({
  data.frame(table_stat,
           Type=c("Frequency",
                  "Central Tendency","Central Tendency","Central Tendency","Central Tendency","Central Tendency",
                  "Accumulation",
                  "Distribution","Distribution","Distribution",
                  "Variability","Variability","Variability",
                  "Distribution","Distribution","Variability","Variability",
                  "Distribution", "Distribution", "Frequency/Distribution"),
           "When/why is this useful?"=
             c("Always useful! Helps understand how much data we are basing the rest of our statistics on",
               "Or the 'arithmetic mean'. 'Average of all of the values' - most easily calculated and commonly used measure of central tendency, but sensitive to outliers and/or non-symmetrical distributions. Forms the basis of most traditional 'frequentist' statistical methodologies and models due to have very friendly mathematical properties.",
               "'The average value' - less sensitive to outliers than the mean, and one that is easier to understand conceptually - half the values are lower than median, and half the values are higher. But a trickier value to model statistically. Often useful to compare to the mean - since very different values for mean and median suggest a distribution that is non-symmetrical and/or has large outliers",
               "The mean value with the top 5% and bottom 5% of values removed, therefore adjusting for the presence of outliers. Can be useful where those outliers are indeed valid observations, but ones that we do not want to focus our analysis on.",
               "The back-transformed average of the logarithm of the values. Useful in specific cases when dealing with exponential distributions - e.g. population growth",
               "The back-transformed average of the reciprocral of the values. Useful in specific cases involving rates or ratios (e.g. the classic school-level maths problem of needing to calcuating the average speed when a driver travels 40km at 40km/h and then 40km at 60km/h. The answer is 48km/h - the harmonic mean; not 50km/h - the arithmetic mean).",
               "Useful if dealing with something that can be summed across the data meaningfully. E.g. the total carbon emissions could be summed here to give us the values for the whole of the UK. But many quantities do not make sense to sum up; and sums are only likely to be useful if we have data covering the whole population rather than a sample of the population.",
               "The lowest observed value - useful for understanding and identifying extremes and/or possible outliers",
            "The highest observed value - useful for understanding and identifying extremes and/or possible outliers",
            "The difference between the min and the max. An extremely sensitive quantity, so only likely to be stable when dealing with the full population rather than a sample.",
            "A measure of variability, which can be interpreted on the same scale as the data itself. If we are in the fortunate case of having something which looks like a normal distribution then 'approximately 95%' of the observed data would be expected to be within +-2SD of the mean. Value is very sensitive to the presence of outliers!",
            "A measure of total variability, the standard deviation squared. Essentially impossible to interpret on it's own, hence why standard deviation is preferred 99% of the time, but has nice statistical properties.",
            "The ratio between the standard deviation and the mean. In certain contexts this can be quite useful to understand - if the %CV is 'small' (e.g. 10%-20%) this suggests a very low amount of variability between values. If it is 'large' (e.g. 200%) this suggests a high amount of variability. Useful for comparison between groups or against known benchmarks. Only valid to consider if all numbers are greater than or equal to zero.",
                   "The point at which 25% of values are lower, and 75% are higher. Useful in combination with the Upper Quartile and Median to understand the shape and variability of the data.",
                "The point at which 75% of values are lower, and 25% are higher. Useful in combination with the Lower Quartile and Median to understand the shape and variability of the data.",
                "The difference between the upper and lower quartiles - a measure of variability since a low value indicates points will be tightly clustered around the median, and high value indicates a wider range of common values. As a measure of variability it is largely unaffected by outliers, unlike the standard deviation/variance.",
                "The median of the deviations between the observations and the median. Like the IQR it is robust in the presence of outliers; unlike the IQR it considers variability across the whole range of data rather than focusing in on the 'middle' 50% of the data.",
                "A measure of the 'shape' of the distribution indicating whether there are 'long tails' - sometimes described fairly abstractly as 'peaked-ness'. A classic normal distribution, a 'bell-shaped curve' has a kurtosis of 0 - negative values indicate something which will have 'smaller' tails (i.e. fewer outlying values), positive values indicate something which will have 'longer' tails (i.e. more outlying values). In practical terms it is almost always less useful than more intuitive methods for determining the presence and impact of outliers, unless you want to look fancy by including some technical terminology that few people understand.",
                "A measure of the 'shape' of the distribution indicating whether the distribution is symmetrical (value>0), 'positively skewed' (value>1) - i.e. has higher density of points below the median but has some values above the median that are very large; or negatively skewed i.e. has higher density of points above the median but some values above the median that are increasingly small. Similar to kurtosis, this is something which is almost always better understood and assessed through more intuitive visual methods.",
                "An intuitive way of understanding the distribution, especially where there may be benchmarks or targets to compare against. For the table above I created an arbitrary target of 100kt which is unlikely to be especially meaningful within this context."
             ),check.names=FALSE)%>% 
    filter(Statistic==input$method) %>%
    gt::gt()
})
```

### Visualisation

All of those numbers can feel a bit overwhelming, and to be honest most
of them are not especially helpful in terms of actually trying to
understand what is going on in my data!

Visual methods for exploration of data are almost always going to be
more useful in uncovering what is going on in our data. As an example of
this there is a famous set of simulated data, the 'Datasaurus dozen',
<https://en.wikipedia.org/wiki/Datasaurus_dozen>, in which the following
13 pairs of variables which all have identical number of observations,
identical means, identical standard deviations and identical
correlations between the sets of variables. And yet they each show
extremely different patterns that would not have been at all apparent
from tables of statistics:

[![Datasauraus dozen:
https://commons.wikimedia.org/wiki/File:EDA_example\_-\_Always_plot_your_data.jpg](https://upload.wikimedia.org/wikipedia/commons/2/29/EDA_example_-_Always_plot_your_data.jpg)](https://upload.wikimedia.org/wikipedia/commons/2/29/EDA_example_-_Always_plot_your_data.jpg)

But, much like with the list of statistics, there are a whole lot of
different options we could use here to visualise our data, and there are
a lot of pros and cons to each of these methods.

In general there is a trade-off we need to make between 'displaying' the
data and 'summarising' the data - methods that only 'display' the data
will be difficult to make inferences from, methods which summarise the
data too much will focus too much on one particular aspect (usually the
average value), and may overlook other key aspects of the distribution
and variability of the variable being plotted.

Take a look at some of the more common different approaches using the
interactive plot below, and with each option think through what these
plots are telling you about the CO2 variable. If there are methods you
are unfamiliar with, then the table in the next section will provide a
brief summary of what they are doing and how they can be interpreted.

```{r v1_ui, context="render",echo=FALSE}

  selectInput("plot_type",label = "Type of Visualisation",
              choices =choices)
  plotlyOutput("plot1")

```

```{r v1_server,context="server"}
  output$plot1 <- renderPlotly({
    
    if(input$plot_type=="boxplot"){
      outs1<-quantile(ghg_2022$Annual.CO2.Total,c(0.25,0.75))+c(-1,1)*diff(quantile(ghg_2022$Annual.CO2.Total,c(0.25,0.75)))*1.5
      outs<-ghg_2022 %>%
        filter(Annual.CO2.Total<outs1[1] | Annual.CO2.Total>outs1[2])
    }
    
( ggplot(ghg_2022)+
      {if(input$plot_type=="point")geom_point(aes(y=Annual.CO2.Total,x=Calendar.Year,county=Local.Authority))}+
      {if(input$plot_type=="caterpillar")geom_point(aes(y=Annual.CO2.Total,x=rank(Annual.CO2.Total),county=Local.Authority))}+
      {if(input$plot_type=="jitter")geom_jitter(aes(y=Annual.CO2.Total,x=Calendar.Year,county=Local.Authority),height=0,width=0.25)}+
      {if(input$plot_type=="errorbar"|input$plot_type=="dynamite")stat_summary(aes(y=Annual.CO2.Total,x=Calendar.Year),fun.data=mean_cl_normal,
                                                                               conf.int=0.95)}+
      # {if(input$plot_type=="dynamite")geom_col(aes(y=mean(Agriculture),x=Calendar.Year))}+
      {if(input$plot_type=="histogram")geom_histogram(aes(x=Annual.CO2.Total))}+
    {if(input$plot_type=="freq_poly")geom_freqpoly(aes(x=Annual.CO2.Total))}+
      {if(input$plot_type=="violin")geom_violin(aes(y=Annual.CO2.Total,x=Calendar.Year,group=Calendar.Year))}+
      {if(input$plot_type=="density")geom_density(aes(x=Annual.CO2.Total))}+
      {if(input$plot_type=="boxplot")geom_boxplot(aes(y=Annual.CO2.Total,outliers=FALSE,
                                                      x=Calendar.Year,group=Calendar.Year,county=Local.Authority,))}+
      {if(input$plot_type=="density"|input$plot_type=="histogram")facet_wrap(~Calendar.Year)}+
      {if(!(input$plot_type=="density"|input$plot_type=="histogram"|input$plot_type=="qq"|input$plot_type=="freq_poly"|
            input$plot_type=="caterpillar"|input$plot_type=="cdf"))scale_x_continuous(limits=c(2021,2023),breaks=2022)}+
      {if(input$plot_type=="dynamite"|input$plot_type=="bar")stat_summary(geom="col",fun="mean",aes(y=Annual.CO2.Total,x=Calendar.Year))}+
      {if(input$plot_type=="cdf")  stat_ecdf(geom = "step",aes(x=Annual.CO2.Total))}+
    {if(input$plot_type=="qq")  stat_qq(aes(sample=Annual.CO2.Total),linewidth=0.5)}+
    {if(input$plot_type=="qq")  stat_qq_line(aes(sample=Annual.CO2.Total),linewidth=0.5,col="red",linetype=2)}+
    {if(input$plot_type=="caterpillar")  labs(x="Rank",y="Annual CO2 Emissions 2022 (kt)")}+
    {if(input$plot_type=="qq")  labs(x="Theoretical Quantile",y="Annual CO2 Emissions 2022 (kt)")}+
    {if(input$plot_type=="cdf")  labs(y="Percentile",x="Annual CO2 Emissions 2022 (kt)")}+
    {if(input$plot_type=="cdf")  scale_y_continuous(labels=scales::percent,breaks=seq(0,1,by=0.2))}+
    {if(input$plot_type=="boxplot")geom_point(aes(y=Annual.CO2.Total,x=Calendar.Year,group=Calendar.Year,county=Local.Authority),
                                              data=outs)}+
    {if(!(input$plot_type=="density"|input$plot_type=="histogram"|input$plot_type=="freq_poly"|input$plot_type=="qq"|
          input$plot_type=="caterpillar"|input$plot_type=="cdf")) labs(y="Annual CO2 Emissions 2022 (kt)",x="") }+
    {if((input$plot_type=="density"|input$plot_type=="histogram"|input$plot_type=="freq_poly")) labs(x="Annual CO2 Emissions 2022 (kt)") }+
    plot_theme+
    ggtitle(paste(names(choices)[choices==input$plot_type]," - Annual CO2 Emissions 2022 (kt)")))  %>% 
      ggplotly()
  })
```

Further explanations about exactly what the plots are presenting, and
why, and some pros and cons can also be found in the table below.

```{r methodp_ui, context="render",echo=FALSE}

  selectInput("method2",label = "Select Visualisation Method",
              choices =c("Point",
"Jittered Plot",
"Boxplot",
"Ridge/Violin Plot",
"Histogram",
"Frequency Polygon",
"Density Plot",
"Caterpillar Plot",
"Bar Chart",
"Dynamite Plot",
"Mean + Error bar",
"Cumulative Density Plot",
"Normal QQ Plot"))
 
  gt::gt_output("table3")


```

```{r methodp_server,context="server"}
output$table3<-render_gt({
  
  data.frame(Plot=c("Point",
"Jittered Plot",
"Boxplot",
"Ridge/Violin Plot",
"Histogram",
"Frequency Polygon",
"Density Plot",
"Caterpillar Plot",
"Bar Chart",
"Dynamite Plot",
"Mean + Error bar",
"Cumulative Density Plot",
"Normal QQ Plot"),
image=choices,
Usages=c("Easy to understand, read off direct data values, and identify outliers",
"Better than a point plot where density or quantity of data is high",
"Visual link between key summary statistics and the data, allows easy comparison across groups in both the averages and the variability",
"Easy comparison of distributions across many groups. Ridge plots and violin plots are effectively the same; only violin plots mirror the distribution shape to be slightly more visually appealing. The shape being plotted is same as in the the density plots, but with the y axis from the density plot removed to allow for multiple 'ridges' or 'violins' to be plotted side by side and compared.",
"Intuitive way to understand the distribution of the data by converting the numeric variable into a categorical variable by creating 'bins' of values and then created a bar chart showing the number of points falling into each of the bins",
"Plots the same information as a histogram, but with a line used to indicate the the frequencies, rather than a bar. This allows for comparison between different groups to be made on the same plot, rather than requiring different plots.",
"A more accurate way to understand the shape of the distribution of the data compared to a histogram or frequency polygon - the density plot smooths out the shape being plotted by the frequency polygon but in a way that is not reliant on an arbitrary choice of how wide to make each 'bin'",
"A good way to identify outliers, and gain better understanding of how different these values are, by plotting the values against the rank of the value - so the highest value will appear on the left hand side, followed by the next highest value and so on. It can also be a really useful method to identify unexpected structure in what was thought to be a numeric value - e.g. if all values are unexpectedly rounded off to the nearest whole number",
"Easy to understand the values being plotted",
"Has the most cool name. Otherwise there are no reasons to ever plot your data like this - please never do it. https://simplystatistics.org/posts/2019-02-21-dynamite-plots-must-die/",
"Allows for easy comparison of the average values across multiple groups by focusing the comparison to a more limited range of values - so can be useful to identify differences when variability is high",
"Easy to read off and link the percentiles across the whole dataset, and to compare across groups when considering the extremes of the distribution (e.g. comparing whether different groups may be at high risk based on looking at differnet cut offs from the 90th/95th/99th percentile). So a good way of summarising a large amount of data efficiently",
"Helps assess if our data is approximately normally distributed in an easier way then other alternatives. The line is the 'theoretical' normal distribution - if the points closely follow the line then we have a distribution that is approximately normal. The QQ plot can also be easily modified to show how data matches to other known distributions (e.g. Posson, Cauchy, Gamma etc)."),
Limitations=c("Cannot infer averages or variability or shape of distribution; points will overlap when there is a large amount or high density of data",
"Solves the issue of overlapping points, but does not solve the other weaknesses listed above",
"Over-simplification of the distribution; outlier detection is relatively arbitrary and often over-interpreted",
"Cannot infer data values, or summary statistics from the plot",
"Shape is easily influenced by the choice of 'bin' width; comparisons can only be made across a small number of groups",
"A little less intuitive than a histogram - with the same limitation of having a shape that is easily influenced by the choice of 'bin' width",
"y axis of 'density' often proves challenging to interpret directly; could be misleading when data sets are small as the volume (or lack of) of data is not apparent from this sort of visualisation",
"Can't easily make comparisons between groups or infer averages or other summary statistics",
"Provides no assessment of variability; makes it difficult to compare between groups where the bar is artificially dragged to start at zero. Should only be used when plotting percentages, counts or sums - should never be used to plot averages",
"All the weaknesses of the 'mean + error bar' plot, without any of the strengths, combined with all of the weaknesses of a bar chart. The lower limit falling under the bar is visually confusing.",
"Error bars based on confidence intervals are based on assumptions, which may or may not be valid, and do not tell us much about the variability in the data - particularly with large datasets",
"Not useful with smaller data sets; usability becomes limited in presence of extreme values",
"... most of the time we don't really care that much about this and this is the only thing it can do")) %>%
      filter(Plot==input$method2) %>%
  gt::gt() %>%
     text_transform(
    locations = cells_body(columns = image),
    fn = function(x) {
      local_image(
        filename = paste0("img/",x,".png"),
        height = 75
      )}
      )%>%
    gt::cols_label(image=" ")
  
})

```

### Outliers

Within some of our plots we could potentially identify three points that
perhaps look a little different to others.

-   There are two points with CO2 emissions that are substantially
    higher than any other local authority - Birmingham and North
    Yorkshire. This can be seen in nearly all of the plots.

-   Within some of the plots, the histogram and the caterpillar plot in
    particular, we may also be able to identify that there is one point
    with very low CO2 emissions relative to the other local
    authorities - the Isles of Scilly

We should take action against outliers if we believe the data to be
incorrect, or the outlier to represent something which comes from a
different process to the rest of our data such that it would be
incomparable. But this should not be a purely data-driven our automated
process - we should not be removing outliers from our data solely
because they look different and their existence is mildly inconvenient
to our analysis. This is likely to influence and bias our results as we
are artificially reducing the variability that exists in the parameters
we are trying to understand. We should be validating our outliers as
genuine observations, and looking for factors that can help to explain
why the results are different.

When considering the presence of outliers there are four key questions
that we need to ask:

-   **Is this a real observation or is this an error?** If it is an
    error than we should be trying to correct the value, and if that is
    not possible then we should exclude the value. Very commonly data
    entry errors may only become apparent when they appear as outliers -
    it is very easy to get decimal places in the wrong position, or add
    too many zeros at the end of large numbers.

-   **If it is real does this observation come from the same underlying
    population as the rest of the observations, and collected using the
    same methodology?** It is possible for data to be 'real' but be
    incomparable to the rest of our observations - we may be "comparing
    apples to oranges" by including them side by side. This is something
    that has been accounted for already in the methodology of this
    dataset - as there are two different estimates for carbon emissions
    presented in the raw data - "total carbon emissions" and "carbon
    emissions under the influence of the local authority". The latter
    estimate removes data that would make comparisons between the local
    authorities "unfair" - for the most part this relates to the
    existence of steelworks and coal-fired power stations, which are
    individually responsible for large amounts of CO2 emissions, but it
    would be unfair to attribute these emissions local authority as they
    would dwarf all other sources of emissions for that authority. Take
    a look for example at the total emissions from Redcar and
    Cleveland - where the total emissions is 6 times larger than the
    "within scope" emissions at the start of the data collection period,
    but by 2022 the two values are fairly similar. This discrepancy is
    entirely related to the Teeside Steelworks,
    <https://en.wikipedia.org/wiki/Teesside_Steelworks>.

```{r}
readRDS("data/outlierplot.RDS") %>% print()
```

-   **If it is real are there other variables that may explain why it is
    so different?** The previous plot also helps illustrate a separate
    point - note the massive drop for the emissions in 2010 and 2011?
    The sudden drop may make us question if this is a mistake, but as is
    outlined in the wiki page - in 2010 and 2011 the Steelworks
    temporarily closed, before being taken over by new owners in 2012.
    It may be that the explanations can be linked to variables within
    our dataset that should then be used within our models. This is
    indeed the case for the three points highlighted in our data -
    Birmingham has an extremely large population, North Yorkshire has a
    large area, The Isles of Scilly has both a low population and a
    small area. These variables will need to be accounted for in our
    final analysis process.

-   **What is the impact of leaving the outlier in our data?** At this
    point in our exploration we should definitely note down the
    potential issue as one to explore later in our analysis to assess
    whether we see the same results with and without the outlier
    included. We would consider this to be an example of a 'sensitivity
    analysis'; if we do see very different results we may choose to
    present both - with caveats included around the inclusion or
    exclusion of potential outliers. An outlier is generally not a
    problem unless it has "high leverage"; we will come back to that
    concept later on in this course.

### Other key characteristics from visual exploration

There is a lot more than just outlier identification that we can see
from this univariate exploration of our CO2 emissions variable!

Some of the key characteristics we should be interpreting from these
plots would be: - The 'average' value of the variable is around 600-700
kt of CO2; although this is the one area when we can see that more
clearly from the summary statistics rather than any of the plots

-   The distribution is 'unimodal'; i.e. in the histogram, density and
    violin points there is one clear peak around which the "average" or
    most common level of CO2 emissions falls

-   The values in the data are truly continuous variables, i.e. they are
    all distinct and not spaced at regular intervals. If data had been
    rounded off, or fell into an ordinal distribution, you would expect
    to see spikes in the shape of the density and violin plots, and
    possibly regularly spaced gaps in the histogram.

-   There is a 'positive' skew in the data - there are an increasingly
    small number of points with increasingly high amount of CO2
    emissions. This is very clearly seen in the histogram, density,
    caterpillar, boxplot & violin plots. In these cases the mean will be
    substantially higher than the median, as we saw in the summary
    statistics table.

-   As a result of this the data is definitely not 'symmetrical' and
    could not be considered to be "normally" distributed. If it was the
    case then the histogram/density plots would fit into a "bell shaped
    curve" and the points in the QQ plot would follow the line

-   There is a large range in the values, which can be seen in most of
    the plots, but the majority of data points fall in a relatively
    narrow range between 200-800 kt of CO2. This aspect can be best seen
    from the relatively small width of the box in the boxplot; the sharp
    peaks in the violin/density/histogram plots; the lack of variability
    in the caterpillar plot; and the rapid vertical ascent in the
    cumulative density plot.

When highlighting some of these key aspects of the plots, you may notice
that most of the different types of plots have been referred to - except
for bar charts, 'dynamite plots' and means with error bars.

As tools for exploratory data visualisation then these are all extremely
limited - bar charts when used for numeric variables generalise a
complex set of data down to a single point and then visually represent
them in a way which is meaningless. There is no reason for a bar chart
representing the emissions to start at 0 - we could choose to draw the
bar starting anywhere, thus making the size of the bar meaningless and
easy to provide misleading plots when we start bringing in additional
variables to compare across.

The use of error bars within exploratory data analysis is also generally
not a good practice - at this stage in the analysis process any
calculation of standard errors or confidence intervals would be based on
naive generic assumptions, that may be wildly incorrect. Using errorbars
is a powerful visual tool to display and communicate the uncertainty in
estimates that we have made after we have concluded our analysis and
modelling which and we will cover in later sessions, but it is not a
good tool to help us understanding the data within this exploratory
phase.

## Exploring One Categorical Variable

Unlike with a numeric variable, where we can be overwhelmed by the
number of options for summary statistics, there are only really two
options for summary statistics when we are dealing with one single
categorical variable - a frequency count or a percentage/proportion.

There is also much less to consider when exploring a single categorical
variable. The three main things we should be considering are:

-   Do we have any errors in our category data entry? "England",
    "england", "ENGLAND" and "England " will all be treated as different
    data levels by most software, so we may need to go back and clean
    our data if we notice that something like this has occurred in our
    data collection process. Not noticing this, and going straight into
    a modelling process is often the source of some fairly major errors
    in analysis.

-   Do we have one 'dominant' category? If the vast majority of our data
    all falls into the same category, it will be difficult to establish
    any relationships between this variable and other variables of
    interest unless we have a very large dataset. In these cases, where
    we would be particularly interested in rare events, we may have been
    better off utilising stratified sampling techniques to over-sample
    rare categories and then utilise weights within our analysis to
    account for this.

-   Do we have several "rare" categories? The opposite scenario is also
    a potential problem - as statistical analysis relies on some degree
    of replication within each group, so having multiple rare categories
    will cause issues when we try to use those variables within an
    analysis. So we may want to consider grouping together categories -
    either on the basis of theoretical similarity or because the number
    of observations is rare. For example - let's say the data available
    covered a slightly wider remit and that within our "Country"
    variable we had one local authority recorded for each of "Isle of
    Man", "Gibraltar", "Guernsey" and "Jersey". It would probably make
    sense to group those entries together into a new category. If the
    entries within the "rare" category did not make intuitive sense to
    link together, we could also group them into an existing variable.
    For example - in the table below if we had had one observation with
    a reduction of "20% or less" and one reduction of "60% or more", it
    would make no sense to group those together. But it would make sense
    to combine the "20% or less" and "30-40%" categories together, and
    create a new label for this; and follow the same process at the
    other end.

```{r sum_table_cat}
ghg_2022 %>%
  group_by(CO2_Reduction_from_2005) %>%
      summarise("Number of Counties"=n(),
                "Percentage of Counties"=scales::percent(n()/nrow(ghg_2022),0.1)) %>%
  gt::gt()
```

### Visualisation

Similarly there are fewer options when considering visualisation of
categorical data. All of them fundamentally distil down to the same idea
of presenting some sort of object that has a size proportional to the
frequency or the percentage.

Although there are perhaps a surprising amount of different options
here:

```{r v2_ui, context="render",echo=FALSE}

  selectInput("plot_type2",label = "Type of Visualisation",
              choices =choices2)
  selectInput("num",label = "Frequencies or Percentages?",
              choices =c("Frequencies","Percentages"))
  
  plotOutput("plot2")
 
  gt::gt_output("table_x")
  
```

```{r v2_server,context="server"}
  
 sum_tab<-   ghg_2022 %>%
  group_by(CO2_Reduction_from_2005) %>%
      summarise(n=n(),
                p=n()/nrow(ghg_2022))

output$plot2 <- renderPlot({
  if(input$plot_type2=="bar")  {
                if(input$num=="Frequencies"){
    p1<-ggplot(sum_tab,aes(x=CO2_Reduction_from_2005,y=n,fill=CO2_Reduction_from_2005))+
          geom_col(show.legend = FALSE,col="black")+
          scale_fill_brewer(palette="Set2")+
          plot_theme
                }
                if(input$num!="Frequencies"){
   p1<- ggplot(sum_tab,aes(x=CO2_Reduction_from_2005,y=p,fill=CO2_Reduction_from_2005))+
          geom_col(show.legend = FALSE,col="black")+
          scale_fill_brewer(palette="Set2")+
        plot_theme
            scale_y_continuous(labels=scales::percent)
                }
  }
        if(input$plot_type2=="count")  {
              if(input$num=="Frequencies"){
    p1<- ggplot(ghg_2022,aes(y=CO2_Reduction_from_2005,x=1))+
          geom_count(show.legend = TRUE,col="black",shape=21,aes(fill=CO2_Reduction_from_2005))+
          scale_fill_brewer(palette="Set2")+
      scale_x_continuous(breaks=NULL)+
      labs(x="")+
     plot_theme       
              }
       if(input$num!="Frequencies"){
    p1<- ggplot(ghg_2022,aes(y=CO2_Reduction_from_2005,x=1,group=1))+
          geom_count(show.legend = TRUE,col="black",shape=21,aes(size = after_stat(prop),fill=CO2_Reduction_from_2005))+
          scale_fill_brewer(palette="Set2")+
      scale_x_continuous(breaks=NULL)+
      labs(x="",size="%")+
        plot_theme+
      scale_size_continuous(labels=scales::percent)
       }
        }
  
      if(input$plot_type2=="stacked_bar")  {
                if(input$num=="Frequencies"){
    p1<- ggplot(sum_tab,aes(x=1,y=n,fill=CO2_Reduction_from_2005))+
          geom_col(show.legend = TRUE,col="black")+
          scale_fill_brewer(palette="Set2")+
        plot_theme
                }
                if(input$num!="Frequencies"){
   p1<-  ggplot(sum_tab,aes(x=1,y=p,fill=CO2_Reduction_from_2005))+
          geom_col(show.legend = TRUE,col="black")+
          scale_fill_brewer(palette="Set2")+
    plot_theme+
            scale_y_continuous(labels=scales::percent)
    
                }
      }
      if(input$plot_type2=="waffle")  {
        if(input$num=="Frequencies"){
    waffle(select(sum_tab,CO2_Reduction_from_2005,n),
       rows=19, size=0.5,title = "CO2 Reduction Across Counties Within The UK",
       xlab="1 Square  = 1 County") %>% print()
        }
                if(input$num!="Frequencies"){
    waffle(select(sum_tab,CO2_Reduction_from_2005,p)%>%mutate(p=round(p*100)),
       rows=10, size=0.5,title = "CO2 Reduction Across Counties Within The UK",
       xlab="1 Square  = 1 % of all counties") %>% print()
        }
      }
      if(input$plot_type2=="pie")  {
             if(input$num=="Frequencies"){   
 p1<-ggplot(sum_tab,aes(x=1,fill=CO2_Reduction_from_2005,y=p))+
  geom_col(col="black",alpha=0.8)+
  coord_polar("y", start=0)+
  scale_fill_brewer(palette="Set2")+
  theme_minimal()+
  scale_x_continuous(breaks=NULL)+
  scale_y_continuous(breaks=seq(0,1,by=0.1),labels=scales::percent)
             }
                if(input$num!="Frequencies"){
 p1<- ggplot(sum_tab,aes(x=1,fill=CO2_Reduction_from_2005,y=n))+
  geom_col(col="black",alpha=0.8)+
  coord_polar("y", start=0)+
  scale_fill_brewer(palette="Set2")+
  theme_minimal()+
  scale_x_continuous(breaks=NULL)
                }
      }
    if(input$plot_type2=="tree_map")  {
    treemap(sum_tab,
            index=c("CO2_Reduction_from_2005"),
            vSize="n",
            type="index",
            palette = "Set2",
            bg.labels=c("white")
            )  %>% print()
    }
    if(input$plot_type2!="tree_map" & input$plot_type2!="waffle")  {
      print(p1)
    }
      })
 
    

output$table_x<-render_gt({
  
  data.frame(Plota=choices2,Plot=names(choices2),

Usages=c("Unlike when using bar charts to display summarised numeric variable there is no data loss by using a bar chart to represent our data, and the     starting point of the bar is clearly defined as 0 when dealing with a frequency or percentage. Side by side bar charts are the simplest to produce, and easiest to read off the data values. But they often result in a lot of 'dead space' on the page, particularly where groups have very different relative frequencies, and when we start to introduce more variables can be hard to compare across groups.",
'Stacked bar charts are better for comparisons, particularly for  the case in which we have ordered categorical data (e.g."None"/"Some"/"A lot"/"All"), but harder to intepret the exact  values. If the categories do not have an inherent order then it can be less useful for comparison.',
"Pie charts have all the negatives of stacked bar charts, as well as being hard to compare across groups like side by side bar charts. And on top of this as humans we are really bad at interpreting the areas presented within pie charts, compared to the rectangles and squares of the other options presented. As a tool for exploring the data, this is not the best, but due to how commonly understood these charts are, it can still be a powerful tool for communicating results.",
"Dot plots use symbols of different sizes to represent the relative counts of frequencies of each group. Somewhat counter intuitively this style of presentation generally  works well in two extremely different scenarios - where there is a high number of groups this can be more visually interpretable than other methods. Where there is a low number of groups this style of presentation is often used with more 'creative' or 'unusual' symbols, rather than just dots, to create something eye-catching. However this is used though it comes with the limitation that it is difficult to read off the exact values for the size of each group.",
"Waffle charts have one square to represent one 'unit' - which can cause problems if we have groups which only take up a fraction of a unit (e.g. a group with a percentage of 0.4% would be dropped completely) - but it is easier to link the visualisation to the actual quantity.",
"Tree maps allocate the area within a rectangle proportionally to each of our categories - there is no way of directly reading off any data values from the visualisation, but it is visually an easy way of comparing the relative sizes of different groups, and it lends itself well to cases where we may have nested or hierachical categorical variables variables (e.g. within our data where we have sub-sectors within sectors within local authorities within years, so could come up with lots of options for nested tree maps using the raw data)")
) %>%
      filter(Plota==input$plot_type2) %>%
    select(-Plota)%>%
  gt::gt() 
  
})

```

With this particular variable, showing the reduction in CO2 since 2005
for each local authority, there is nothing that we should be too worried
about. There are no issues with labelling of the categories; although
the 40-50% category is the most frequent it is not approaching a level
of dominance that would indicate it might cause problems for our
analysis; and there are at least 20 counties in each of our four
categories.

That final point matches perfectly a somewhat arbitrary rule of thumb
that is used when determining if a category is 'too rare' - where we
would be looking for at least 20 observations per group to be wanting to
include it a statistical model. In reality, setting a minimum number of
observations per group required for use in modelling is a little more
complex than this, depending on the relationships between this variable
and the rest of the data, and what modelling techniques are to be used,
but in general it is not a bad rule of thumb to have in mind at this
stage.

## Two Variables

A lot of summary statistics and visualisations to identify relationships
between two or more variables build upon those we have covered so far.

When comparing how two variables relate to each other it is likely we
may have some hypotheses or questions already that we want to explore
(e.g. how have CO2 levels changed over time? Are CO2 levels more
impacted by population or area? Which sectors have made the biggest
changes in emissions over time?). But sometimes we may come in with no
real preconceptions around how the relationships in our data may look.

### Comparing Numeric vs Categorical

All of the summary statistics covered previously would work in exactly
the same way as we saw already, except we now have multiple groups to
compare our statistics across. Even having some degree of comparison
across groups defined by categorical variables can help to make many of
these statistics a bit more useful! Let's consider splitting them by
country:

```{r sum_table grp}
ghg_2022 %>%
  group_by(Country) %>%
      summarise("Number of Counties"=n(),
                Mean=mean(Annual.CO2.Total)%>%round(1),
                Median=median(Annual.CO2.Total)%>%round(1),
                "Trimmed mean"=mean(Annual.CO2.Total,trim = 0.05)%>%round(1),
                 "Geometric mean"=psych::geometric.mean(Annual.CO2.Total)%>%round(1),
                "Harmonic mean"=psych::harmonic.mean(Annual.CO2.Total)%>%round(1),
                 Sum=sum(Annual.CO2.Total)%>%round(),
                "Minimum"=min(Annual.CO2.Total)%>%round(1),
                "Maximum"=max(Annual.CO2.Total)%>%round(1),
                "Range"=(max(Annual.CO2.Total)-min(Annual.CO2.Total))%>%round(1),
                "Standard Deviation"=sd(Annual.CO2.Total)%>%round(1),
                "Variance"=var(Annual.CO2.Total)%>%round(),
                "Coefficient of Variation"=scales::percent(sd(Annual.CO2.Total)/mean(Annual.CO2.Total)),
                "Lower Quartile"=quantile(Annual.CO2.Total,0.25)%>%round(1),
                "Upper Quartile"=quantile(Annual.CO2.Total,0.75)%>%round(1),
                "Inter Quartile Range"=diff(quantile(Annual.CO2.Total,c(0.25,0.75)))%>%round(1),
                "Median Absolute Deviation"=mad(Annual.CO2.Total)%>%round(1),
                "Kurtosis"=agricolae::kurtosis(Annual.CO2.Total)%>%round(2),
                "Skewness"=agricolae::skewness(Annual.CO2.Total)%>%round(2),
                "% of Values Over Threshold (100kt)"=scales::percent(mean(Annual.CO2.Total>1000)),
                )%>%
  gather(Statistic,Value,-Country ,factor_key = TRUE) %>%
  spread(Country ,Value) %>%
  gt::gt() 
  
```

But of course, the same limitations apply - and we will nearly always
get a better understanding from comparing groups visually.

Instead of looking at one single box in a boxplot we are now looking at
two (or more) boxes side by side.

Many of the methods covered previously for visualising one variable in
fact become a lot more useful in this context, of comparing one variable
across multiple groups. But not all of the methods covered are so
suitable for comparison - qq plots for example may be produced for
different groups, and we may be interested in understanding the results
within each group, but there is no real benefit in visually comparing
the relative 'normal-ness' of different groups. In these cases we may
chose to plot different groups in different panels, rather than overlay
the groups within the same plot.

```{r v3_ui, context="render",echo=FALSE}

  selectInput("plot_type3",label = "Type of Visualisation",
              choices =choices)
  plotlyOutput("plot3")

```

```{r v3_server,context="server"}
  output$plot3 <- renderPlotly({
    
    if(input$plot_type3=="boxplot"){
      outs1<-quantile(ghg_2022$Annual.CO2.Total,c(0.25,0.75))+c(-1,1)*diff(quantile(ghg_2022$Annual.CO2.Total,c(0.25,0.75)))*1.5
      outs<-ghg_2022 %>%
        filter(Annual.CO2.Total<outs1[1] | Annual.CO2.Total>outs1[2])
    }
    
( ggplot(ghg_2022)+
      {if(input$plot_type3=="point")geom_point(aes(y=Annual.CO2.Total,x=Country,colour = Country,county=Local.Authority))}+
      {if(input$plot_type3=="caterpillar")geom_point(aes(y=Annual.CO2.Total,colour = Country,x=rank(Annual.CO2.Total),county=Local.Authority))}+
      {if(input$plot_type3=="jitter")geom_jitter(aes(y=Annual.CO2.Total,colour = Country,x=Country,county=Local.Authority),height=0,width=0.25)}+
      {if(input$plot_type3=="errorbar"|input$plot_type3=="dynamite")stat_summary(aes(y=Annual.CO2.Total,colour = Country,x=Country),fun.data=mean_cl_normal,
                                                                               conf.int=0.95)}+
      # {if(input$plot_type=="dynamite")geom_col(aes(y=mean(Agriculture),x=Calendar.Year))}+
      {if(input$plot_type3=="histogram")geom_histogram(aes(x=Annual.CO2.Total,colour = Country))}+
    {if(input$plot_type3=="freq_poly")geom_freqpoly(aes(x=Annual.CO2.Total,colour = Country))}+
      {if(input$plot_type3=="violin")geom_violin(aes(y=Annual.CO2.Total,x=Country,fill = Country,group=Country))}+
      {if(input$plot_type3=="density")geom_density(aes(x=Annual.CO2.Total,col = Country))}+
      {if(input$plot_type3=="boxplot")geom_boxplot(aes(y=Annual.CO2.Total,outliers=FALSE,fill = Country,
                                                      x=Country,group=Country,county=Local.Authority,))}+
      {if(input$plot_type3=="histogram"|input$plot_type3=="caterpillar"|
          input$plot_type3=="histogram"|input$plot_type3=="qq")facet_wrap(~Country)}+

      {if(input$plot_type3=="dynamite"|input$plot_type3=="bar")stat_summary(geom="col",fun="mean",aes(y=Annual.CO2.Total,x=Country,fill = Country))}+
      {if(input$plot_type3=="cdf")  stat_ecdf(geom = "step",aes(x=Annual.CO2.Total,colour = Country))}+
    {if(input$plot_type3=="qq")  stat_qq(aes(sample=Annual.CO2.Total,colour = Country),linewidth=0.5)}+
    {if(input$plot_type3=="qq")  stat_qq_line(aes(sample=Annual.CO2.Total),linewidth=0.5,col="black",linetype=2)}+
    {if(input$plot_type3=="caterpillar")  labs(x="Rank",y="Annual CO2 Emissions 2022 (kt)")}+
    {if(input$plot_type3=="qq")  labs(x="Theoretical Quantile",y="Annual CO2 Emissions 2022 (kt)")}+
    {if(input$plot_type3=="cdf")  labs(y="Percentile",x="Annual CO2 Emissions 2022 (kt)")}+
    {if(input$plot_type3=="cdf")  scale_y_continuous(labels=scales::percent,breaks=seq(0,1,by=0.2))}+
    {if(input$plot_type3=="boxplot")geom_point(aes(y=Annual.CO2.Total,x=Country,group=Country,county=Local.Authority,colour = Country),
                                              data=outs)}+
    {if(!(input$plot_type3=="histogram"|input$plot_type3=="qq"|
          input$plot_type3=="caterpillar"|input$plot_type3=="cdf")) labs(y="Annual CO2 Emissions 2022 (kt)",x="") }+
    {if((input$plot_type3=="histogram")) labs(x="Annual CO2 Emissions 2022 (kt)") }+
   plot_theme+
    ggtitle(paste(names(choices)[choices==input$plot_type3]," - Annual CO2 Emissions 2022 (kt)")))  %>% 
      ggplotly()
  })
```

### Categorical vs. Categorical

When exploring the relationship between two (or more) categorical
variables there is just one additional source of complexity to consider
when summarising the results - what percentages are the most meaningful
or useful to calculate?

```{r v7_ui, context="render",echo=FALSE}

  selectInput("percent_type",label="Type of Percentage",choices = c("row","column"="col","overall"="all"),
             )

  gt_output("crosstab")

```

```{r sum_table_cat7, context="server",echo=FALSE}


output$crosstab<-render_gt({
  ghg_2022 %>%
  tabyl(CO2_Reduction_from_2005,Country) %>%
      adorn_percentages(denominator = input$percent_type)%>%
      adorn_pct_formatting() %>%
      adorn_ns(format_func = function(x)paste("n =",x)) %>%
  gt()
})

```

In the table above which sort of percentage is likely to make the most
sense?

-   (within) row percentages: '92% **of the counties who reduced CO2 by
    50% or more** were in England'

-   (within) column percentages : '11% **of the counties in England**
    reduced CO2 by 50%'

-   overall percentages : 9% of **all** counties reduced CO2 by 50% or
    more **and** were in England

Sometimes multiple versions of the percentages may be of interest;
although it is usually the case that there will be one which is clearly
the most appropriate option to be using. In this case, and in cases were
there is a clear distinction between an 'outcome' variable (like CO2
emissions) and an 'explanatory' variable (like country), there is a
clear answer as to which would be the most sensible to consider.

So rhe 'column' percentages from the table above will be the most useful
here - being able to compare the % of counties falling into each of the
CO2 reduction categories by country tells us some interesting things
about differences between the countries of the UK.

Looking at the 'row' percentages here does not tell us much at all. Out
of all of the counties in the UK 296/361 are in England. Therefore it
makes it close to impossible to work out if the distribution across the
four counties is really that different when comparing the percentages
within each reduction category.

The overall percentages do not make sense at all here. They would only
really make sense to compute where we were interested in combining these
two variables together to form a new composite variable of the
interaction between the categories of these variables.

Remember that the "row" and the "column" percentages refer only to the
orientation of the variables in the table - whether the percentages
within each row add up to 100%, or whether the percentages within each
column add up to 100% . The 'useful' percentages would become the "row"
percentages should we have decided to put the countries as the variable
making up the rows, and the CO2 reduction categories as the variable
making up the "columns".

A similar choice is required when producing visualisations, previously
it did not matter whether we plotted frequencies or percentages as
visually this would end in the same result.

When dealing with two (or more) variables then plotting the frequencies
is more or less equivalent to plotting the "overall" percentages - where
it may help to partition the overall dataset into subgroups based on the
two variables, but it does not help us understand the relationship
between those variables.

The waffle plot or tree map approach is one that only really makes sense
when dealing with the "overall" percentages, or frequencies, as it will
be visually misleading to present within group percentages. You can see
an example of this in the interactive menu below where the presentation
of each country taking the same size does not intuitively make sense in
the same way as the other visualisation options.

Stacked bar charts come into their own here in this case where we are
comparing an ordinal categorical variable, as it lets us accumulate
between adjacent categories from the top or the bottom of the stack so
that we can start visually comparing, for example, those with "40% or
lower" reduction emissions - not just those with "30% or lower". Making
visual comparisons between pie charts is something which is a
challenge - there is no way to do this intuitively.

```{r v22_ui, context="render",echo=FALSE}

  selectInput("cat_plot2",label = "Type of Visualisation",
              choices =choices2[-4])
  selectInput("num_t2",label = "Frequencies or Percentages?",
              choices =c("Frequencies","Percentages"))
  
  plotOutput("catplot2")


```

```{r v22_server,context="server"}
  
 sum_tab2<-   ghg_2022 %>%
  group_by(Country,CO2_Reduction_from_2005) %>%
      summarise(n=n()) %>%
  mutate(p=n/sum(n))

output$catplot2 <- renderPlot({
  if(input$cat_plot2=="bar")  {
                if(input$num_t2=="Frequencies"){
    p1<-ggplot(sum_tab2,aes(x=Country,y=n,fill=CO2_Reduction_from_2005))+
          geom_col(show.legend = FALSE,col="black",position="dodge")+
          scale_fill_brewer(palette="Set2")+
          plot_theme
                }
                if(input$num_t2!="Frequencies"){
   p1<- ggplot(sum_tab2,aes(x=Country,y=p,fill=CO2_Reduction_from_2005))+
          geom_col(show.legend = FALSE,col="black",position="dodge")+
          scale_fill_brewer(palette="Set2")+
         plot_theme+
            scale_y_continuous(labels=scales::percent)
                }
      }
      if(input$cat_plot2=="stacked_bar")  {
                if(input$num_t2=="Frequencies"){
    p1<- ggplot(sum_tab2,aes(x=Country,y=n,fill=CO2_Reduction_from_2005))+
          geom_col(show.legend = TRUE,col="black")+
          scale_fill_brewer(palette="Set2")+
     plot_theme
                }
                if(input$num_t2!="Frequencies"){
   p1<-  ggplot(sum_tab2,aes(x=Country,y=p,fill=CO2_Reduction_from_2005))+
          geom_col(show.legend = TRUE,col="black")+
          scale_fill_brewer(palette="Set2")+
          plot_theme+
            scale_y_continuous(labels=scales::percent)
    
                }
      }

      if(input$cat_plot2=="pie")  {
             if(input$num_t2=="Frequencies"){   
 p1<-ggplot(sum_tab2,aes(x=1,fill=CO2_Reduction_from_2005,y=n))+
  geom_col(col="black",alpha=0.8)+
  coord_polar("y", start=0)+
  scale_fill_brewer(palette="Set2")+
  theme_minimal()+
  scale_x_continuous(breaks=NULL)+
  scale_y_continuous(breaks=seq(0,1,by=0.1),labels=scales::percent)+
   facet_wrap(~Country,scales="free")
             }
                if(input$num_t2!="Frequencies"){
 p1<- ggplot(sum_tab2,aes(x=1,fill=CO2_Reduction_from_2005,y=p))+
  geom_col(col="black",alpha=0.8)+
  coord_polar("y", start=0)+
  scale_fill_brewer(palette="Set2")+
  theme_minimal()+
  scale_x_continuous(breaks=NULL)+
   facet_wrap(~Country,scales="free")
                }
      }
    if(input$cat_plot2=="tree_map")  {
    treemap(sum_tab2,
            index=c("Country","CO2_Reduction_from_2005"),
            vSize=ifelse(input$num_t2=="Frequencies","n","p"),
            type="index",
            vColor="CO2_Reduction_from_2005",
            palette = "Set2",
            bg.labels=c("white")
            )  %>% print()
    }
     if(input$cat_plot2=="count")  {
              if(input$num_t2=="Frequencies"){
    p1<- ggplot(ghg_2022,aes(y=CO2_Reduction_from_2005,x=Country))+
          geom_count(show.legend = TRUE,col="black",shape=21,aes(fill=CO2_Reduction_from_2005))+
          scale_fill_brewer(palette="Set2")+
         labs(x="")+
     plot_theme       
              }
       if(input$num_t2!="Frequencies"){
    p1<- ggplot(ghg_2022,aes(y=CO2_Reduction_from_2005,x=Country,group=Country))+
          geom_count(show.legend = TRUE,col="black",shape=21,aes(size = after_stat(prop),fill=CO2_Reduction_from_2005))+
          scale_fill_brewer(palette="Set2")+
      labs(x="",size="%")+
        plot_theme+
      scale_size_continuous(labels=scales::percent)
       }
        }
    if(input$cat_plot2!="tree_map" & input$cat_plot2!="waffle")  {
      print(p1)
    }
      })
 
    

```

## Comparing Two Numeric Variables

When considering the relationship between two numeric variables there is
a much more straightforward choice to be made as there is only really
one immediately obvious exploratory plot to explore - a scatter plot.

Let's consider the relationship between the CO2 emissions and the
population of the county.

```{r}
(ggplot(data=ghg_2022,aes(x=Mid.year.Population..thousands.,y=Annual.CO2.Total,county=Local.Authority))+
  geom_point()+plot_theme) %>% ggplotly()
```

The conventional wisdom would be to place the variable we might consider
to be our "outcome" or "response" variable on the y axis, and the
variable we would consider to be our "explanatory" or "predictor"
variable on the x axis, to help visually link the direction of the
relationship between variables.

We can immediately see what appears to be a very strong looking
correlation between the two variables - as population increases so does
the annual CO2 emissions.

In this case the direction is relatively clear - we would be considering
emissions as the outcome variable and population as the explanatory
variable - we would expect the carbon emissions to increase as the
population increases, rather than vice versa. There may be cases where
there is really no clear choice for which variable is the "outcome" -
when considering the link between the population and area variables for
example - and in this case it does not matter so much which orientation
we would choose to place our two variables.

We may want to add trend lines to the plot, which will help us
understand the shape of the relationship. At this point in the
exploration it is based to fit a smoothed line onto our plot - this will
guide us as to the shape of the trend. If the smoothed curve looks close
to being a straight line, then a simple linear relationship may be
appropriate. Or alternatively the smooth trend line may suggest the
pattern follows a known consistent curve than could be modelled for -
showing exponential increase, or a quadratic relationship, or an s-shape
logistic curve. These consistent curves have the advantage within
statistical modelling that they can be 'parameterised', and therefore
used to form the basis of an equation that underlies the relationship
and can help us link the data to a theoretical relationship. Smoothed
curves do not have this advantage, although they can instead be used in
more predictive applications of modelling, rather than applications
focused on interpretation of the relationships.

In the plot below you can choose to add a trend line based on a simple
straight line relationship, a smoothed curve using either a generalised
additive model ('gam'), or a moving average. The latter two options will
broadly produce the same type of interpolated line - the moving average
model is a little more conservative, and a better choice for smaller
datasets, whilst the 'gam' method would be more likely to identify
localised patterns within larger datasets (\>1000 observations) but may
be a little too easily influenced by individual observations in smaller
datasets.

```{r trd_in,context="render"}
selectInput("trendline","Trendline?",choices=c("None","Linear"="lm","Smooth"="gam","Moving Average"="loess"))
plotlyOutput("trend")
```

```{r trd_out,context="server"}
output$trend<-renderPlotly({
  
  (ggplot(data=ghg_2022,aes(x=Mid.year.Population..thousands.,y=Annual.CO2.Total))+
  geom_point(aes(county=Local.Authority))+plot_theme+
 {if(input$trendline!="None") geom_smooth(method=input$trendline)}
  ) %>% ggplotly()
})
```

In this case it seems as if the relationship between our variables is
indeed very close to a simple linear straight line: as the population of
a county increases by (x) thousand the CO2 emissions increase by (y)
kilo tonnes, with very minimal evidence of curvature or deviation from
this general trend.

### Correlation

You are probably also familiar with the concept of a 'correlation
coefficient' which is a summary statistic to assess if two numeric
variables are positively correlated (i.e. as one variable increases the
other also increases; a positive value for the correlation coefficient),
negatively correlated (i.e. as one variable increases the other
decreases; a negative value for the correlation coefficient), or
uncorrelated (no relationship between the variables a value close to
zero).

The further away the coefficient is from zero, the stronger relationship
is - it is often common to see a correlation matrix, which will assess
all pairwise correlations between numeric variables in your dataset:

```{r}
ghg_2022 %>%
  ungroup() %>%
  select(Area..km2.,Mid.year.Population..thousands.,Annual.CO2.Total) %>%
      cor() %>%
        data.frame() %>%
        gt(rownames_to_stub = TRUE)%>%
          fmt_number()
```

This matrix shows us that:

-   All pairwise correlations between our three variables (area,
    population, and CO2 emissions) are positive
-   The correlation between CO2 emissions and population is very
    strong - generally this would be considered a correlation of 0.7 or
    higher
-   The correlation between CO2 emissions and area is moderate -
    generally as a rule of thumb this could be considered a correlation
    of 0.25-0.5
-   The correlation between population and area is moderate - generally
    as a rule of thumb this could be considered a correlation of
    0.25-0.5

Also notice that the correlation between a variable and itself is equal
to 1; area is of course perfectly correlated to area! And that the "top
right" and "bottom left" corners of the matrix are identical - the
correlation between area and population, is identical to the correlation
between population and area. So despite being a table of 9 numbers,
there are only 3 that are of any interest to us.

Like with the scatter plot, this is the immediately obvious choice for a
summary statistic when considering two numeric variables, but therere
are lots of different variations of the correlation coefficient. The
"default" correlation method that you are likely familiar with, and has
been applied in the plot above, isthe "Pearson" method. Some of the most
common alternatives, "Spearman" & "Kendall", are summarised in the table
below with the rationale for when they might be applied.

```{r}

data.frame("Correlation Type"=c("Pearson (rho)","Spearman (rho)","Kendall (tau)"),
           "Variables"=c("Continuous numeric data only","Continous or ordinal data","Continous or ordinal data"),
           "Relationship Type"=c("Linear only","Monotonic (i.e. can be curved providing curve is either strictly increasing or strictly decreasing","Monotonic (i.e. can be curved providing curve is either strictly increasing or strictly decreasing"),
           "Pros"=c("Highest statistical 'power', easiest to interpret and links directly to other statistical concepts (e.g. R2)","Robust to either the presence or absence from linearity, if conditions for Pearson are satisfied then Spearman is likely to be a very similar value, so no net loss of choosing this option","Most robust to 'imperfections'"),
           "Cons"=c("Very sensitive to presence of outliers","Harder to interpret directly, or to link to other methods. Over-estimates correlation in presence of ties.","Will always show the 'weakest' correlation of the three methods even where conditions are satisfied")) %>% gt()


```

In statistics we often see this trade-off in choice of methodologies
between 'power' and 'robust-ness' - the "Kendall" method for correlation
is perhaps the most robust, but the least powerful - it will always be
the smallest of the 3 measures. Whilst the "Pearson" method is the most
powerful but least robust, it is highly sensitive to outliers in small
data sets or to non linear trends. It is often incorrectly claimed that
"Pearson" correlation also requires the variables to be normally
distributed; but this is not actually true.

If the conditions for the Pearson method are satisfied (a consistent
linear relationship, with no, or very few, data points sharing the exact
same value, then it is likely that this version of the correlation
coefficient will be the largest, providing the strongest evidence of
correlation, but that all three methods will give broadly similar
results.

You can explore how the choice of correlation method affects the values
in the interactive table below:

```{r cor_in,context="render"}
  selectInput("cor_m","Correlation Method",choices =c("pearson","spearman","kendall"),selected = "pearson")
gt_output("cor_tab1")
```

```{r cor_tab,context="server"}
output$cor_tab1<-render_gt({ghg_2022 %>%
  ungroup() %>%
  select(Area..km2.,Mid.year.Population..thousands.,Annual.CO2.Total) %>%
      cor(method=input$cor_m) %>%
        data.frame() %>%
        gt(rownames_to_stub = TRUE,caption=paste("Correlation using",input$cor_m,"method"))%>%
          fmt_number()})
```

Remember that these pairwise correlations may be hiding a more complex
relationship - either one that is not monotonic or one that may be
moderated, or influenced, by other variables. Take the relationship
between area and population. The Spearman and Kendall coefficients for
this relationship are both very close to zero, and in fact negative.

And again remember that all of these methods of correlation are useless
if the relationship between the two variables does not strictly increase
or decrease - you can go back to the previous section, and remind
yourself of the "datasaurus" dozen, where all 13 of the plots showed the
same lack of correlation between the x and y variables, despite the two
variables being very clearly related in many of the plots.

So in practice, looking at the relationship through the scatter plot is
a vastly superior method of understanding your data. And this comes with
an additional advantage of being easy to extend scatter plots further to
show beyond just a simple x/y relationship.

### Going further with scatter plots

Beyond just establishing the potential existence and shape of a trends,
there are many further questions we can visually explore with scatter
plots:

-   Is the variability away from the trend consistent or does it vary
    (e.g. is there higher variation when we have higher values?)
-   Is the relationship consistent across different groups?
-   Are there values which do not fit into the trend? Are these
    mistakes, or are there other factors which may explain why they are
    different?
-   If there are other factors, can we start to bring in additional
    variables to explore the relationship in 3 (or more) dimensions

From the plots we have made so far there are some questions or next
steps that we might take from our original plot, once we have
acknowledged the clear strong trend that exists:

-   The density of points is very high in the bottom left of the plot -
    there are lots of counties with relatively small populations and
    only a small number of counties with larger populations than others.
    Birmingham, North Yorkshire and Leeds all have points which are very
    distinct from the others. At the same time the deviation from the
    trend looks like it seems to get larger as we increase the
    population. Look at the point for North Yorkshire which appears to
    be something of an outlier compared to the trend. If you were to
    compare it to Glasgow, for example, both have very similar
    populations, but the emissions for North Yorkshire are around 3500
    kt/CO2 and for Glasgow around 1950 kt/CO2. The emissions from North
    Yorkshire are similar to that of Birmingham, where the population is
    almost twice as large.

We saw this sort of trend a little with the CO2 and population variable
but it is extremely pronounced with the area variable.

```{r}
ghg_2022 %>%
  ggplot(aes(x=Area..km2.))+
  geom_histogram()+
  plot_theme
```

For the area variable this is the exact sort of pattern where we would
want to consider a log transformation for population, or potentially for
both of the variables. This will help us to see inside the dense cloud
of points currently at the bottom left rather than only focusing on the
small number of points with high values. There are other transformations
beyond log transformations that we could choose to explore - and there
will be particular scientific contexts where different transformations
will be scientifically coherent - but it is usually a good idea to
consider whether that transformation can be interpreted coherently
rather than just being a 'good fit' for the data. Log transformations
are nice in this way as they can, with a little bit of additional
thought, usually be interpreted.

Considering a log transformation for population only, the question the
plot is assessing changes from whether "as the size of the population by
increases by (x) people does the CO2 emissions increase by (y) kt?" on
the original scale, to a new question of "as the size of the population
**doubles in size** does the CO2 emissions increase by (y) kt?".
Remember though that this sort of transformation will only work when the
variable is always larger than zero, as we cannot think about logarithms
and rates of change when the values can plausibly be negative or equal
to zero.

For area, it is of clear benefit to perform a transformation, but for
population or CO2 emissions it is less clear cut. So it may be something
to consider in our formal analysis where it may or may not be a useful
transformation to help us model the relationship better.

-   Having uniformly sized and styled points for all counties may be a
    little limited. We could bring in additional variables to map to the
    size and colour of the points - perhaps different colours for the
    four nations of the UK, and different sizes based on the area of the
    county?

You can explore these options, and more, by adjusting the selections in
the interactive plot below.

```{r v5_ui, context="render",echo=FALSE}
fluidRow(column(
  selectInput("x1","Variable for x axis",choices =c("Population"="Mid.year.Population..thousands.","Area"="Area..km2.")),
  selectInput("x_t","Transform x axis?",choices =c("No","log"),selected = "log"),
  selectInput("y_t","Transform y axis?",choices =c("No","log"),selected = "log"),width=6),
column(
  selectInput("psize","Point Size",
              choices =c("Fixed","Population"="Mid.year.Population..thousands.","Area"="Area..km2."),
              selected = "Area..km2."),
  selectInput("pcol","Point Colour?",choices =c("Fixed","Country","Region","Reduction since 2005"="CO2_Reduction_from_2005"),
              selected="Country"),width=6),
)
  plotlyOutput("plot_scatter")

```

```{r v5_server,context="server"}
  output$plot_scatter <- renderPlotly({
    ghg_2022<-data.frame(ghg_2022)
     ghg_2022$y<-ghg_2022$Annual.CO2.Total
    ghg_2022$No<- ghg_2022$Fixed<-1
    ghg_2022$x<-ghg_2022[,input$x1]
      ghg_2022$size<-ghg_2022[,input$psize]
        ghg_2022$colour<-ghg_2022[,input$pcol]
        
( ggplot(ghg_2022,aes(y=y,x=x))+
    geom_point(aes(colour=colour,size=size,county=Local.Authority),alpha=0.6)+
    {if(input$x_t=="log") scale_x_log10()}+
    {if(input$y_t=="log") scale_y_log10()}+
    scale_size_continuous(range = c(0.5,3.5))+
    theme_light()+
    labs(x=input$x1,y=paste("Annual CO2 Emissions",ifelse(input$y_t=="No","",input$y_t)))+
    ggtitle(paste("Annual CO2 Emissions 2022 (kt) by",input$x1)) ) %>% 
      ggplotly()
  })
```

By trying different options in our exploration, this will will help us
to understand even more about our data!

-   The large deviation in the trend for North Yorkshire does indeed
    look be linked to its very large area as we suspected earlier
-   The biggest outlier when assessed on the log scale, may in fact be
    the City of London, which has a very low population but a high level
    of emissions. This itself is likely to have some clear
    explanations - the number of people working within the City of
    London is extremely high, but the number of permanent residents is
    low. There may exist some variable to quantify this, based on
    economic activity, but this is not included within the dataset that
    we have. Sometimes deviations we see from the trends would require
    additional data to confirm.

### Exploring Trends Over Time

'Time' is perhaps a special case of a numeric variable - and in fact in
different sections of this tutorial we have in fact considered "year" as
a numeric or sometimes as a categorical variable. In general if we are
dealing with more than a small number of numeric observations we would
want to consider time broadly more like a numeric variable. Beyond this
there are also a few additional things to consider when dealing with
time, and particularly when dealing with data recorded on a more regular
basis than once per year:

-   Conventional wisdom would suggest always placing time under the y
    axis and plotting trends over time as a line connecting the
    observations, rather than plotting points for each observation
-   When dealing with data recorded at very frequent intervals plotting
    every single recording is likely to give a lot of noise; some
    decisions would have to be made about what level to aggregate the
    data to - every minute? every hour? every day?. It is worth
    exploring the data and producing summaries at all of these levels,
    and any other relevant ones, to better understand the periodicity
    and auto-correlation in the trends over time. Periodicity refers to
    consistently repeating patterns - e.g. temperatures getting warmer
    in the day and then colder in the night; while auto-correlation
    refers to how correlated adjacent observations would be - e.g. the
    temperature now is likely to be extremely highly correlated to the
    temperature 10 seconds ago; somewhat correlated to the temperature
    in 24 hours time; and, after accounting for the seasonal effects
    (which are part of the periodicity), very weakly or not at all
    correlated to the temperature in one months time. Plotting the data
    at different time resolutions lets us explore these components of
    time series, which are very important to understand before moving to
    formal time series modelling.
-   When joining together data to form a line over time there is a
    visual question of whether to join each data point directly
    together, or whether to create steps, like a staircase. Often this
    will depend on whether you have regularly spaced data or 'event'
    data. If the change in the variable happen gradually between the
    time points, like may be expected where there are regular intervals,
    then a straight line would make more sense. If the change in the
    value happens at the exact time point being recorded then it would
    not make sense to interpolate between the points, so drawing 'steps'
    may make more sense.
-   The 'area under the curve' is often a useful summary statistic to
    consider when dealing with trends over time. Within different
    contexts for different variables being plotted against time it can
    have different direct interpretations, but in general it is a
    measure of intensity. A variable which quickly rises to a high peak
    value, and then quickly falls back down, will have a lower "AUC"
    compared a variable which may reach a lower peak but remains around
    that peak for a longer period of time.

Let's take a look at a slightly different extract of the data we have
seen already - this time looking at the total CO2 emissions per country
over the full time period of the data 2005-2022.

```{r}
ghg_yearly<-ghg_allyears %>%
  group_by(Country,Calendar.Year) %>%
    summarise(Annual.CO2.Total=sum(Annual.CO2.Total),
              Mid.year.Population..thousands.=sum(Mid.year.Population..thousands.),
              Area..km2.=sum(Area..km2.))


DT::datatable(ghg_yearly)
```

```{r time_ui1, context="render",echo=FALSE}


  selectInput("plot_type_time1",label = "Type of Visualisation",
              choices = c("point",
                          "line",
                          "step"              ))  
  plotlyOutput("time1")

```

```{r time_server1,context="server"}

output$time1 <- renderPlotly({
    ghg_yearly<-ghg_allyears %>%
  group_by(Country,Calendar.Year) %>%
    summarise(Annual.CO2.Total=sum(Annual.CO2.Total),
              Mid.year.Population..thousands.=sum(Mid.year.Population..thousands.),
              Area..km2.=sum(Area..km2.))
   
    
   ( ggplot(ghg_yearly)+
      {if(input$plot_type_time1=="point")geom_point(aes(y=Annual.CO2.Total,x=Calendar.Year,colour=Country))}+
      {if(input$plot_type_time1=="line")geom_line(aes(y=Annual.CO2.Total,x=Calendar.Year,colour=Country))}+
      {if(input$plot_type_time1=="step")geom_step(aes(y=Annual.CO2.Total,x=Calendar.Year,colour=Country))}+
       theme_light()+
       scale_x_continuous(breaks=2005:2022)) %>% ggplotly()
    
  })

```

### Derivations

As well as the choice of visual method, there are further data
manipulations, derivations and transformations which will help us to
understand the trends better here. Specifically to time series we could
consider plotting the change over time since baseline, in 'real terms'
or in 'percentage terms' instead of the actual values on the y axis.
Plotting this in percentage terms is a good option here to make a fair
comparison in the trend across the four countries, at the trade-off of
being able to easily read off from the plot the actual CO2 values.

The log or square root transformations are not especially useful in this
plot! But adjusting the emissions per capita, or per area, are perhaps
better options to help make a much fairer comparison across the 4
groups, which have vastly different areas and populations. This requires
us to derive new variables, in this case by dividing the emissions by
the population or area.

And when looking "per capita" we can see very similar trends across the
four countries, with the line almost following in parallel to each
other, all generally decreasing over time. When looking at the
percentage change over time the trend is very similar for three of the
four countries - with the lines not just parallel but in fact almost
identical over time. The difference is from Northern Ireland - where the
percentage reduction over time was lower, with emissions in fact
increasing for the first few years of this data.

```{r time_ui2, context="render",echo=FALSE}


  selectInput("plot_type_time",label = "Type of Visualisation",
              choices = c("line",
                          "step"              ))
  
  selectInput("trans_t",label = "Transform Variable",
              choices = c("None"="none",
                          "Logarithmic"="log",
                          "Square Root"="sqrt",
                          "Per Capita"="capita",
                          "Per Area"="area",
                          "Change from baseline"="rebase",
                          "% change from baseline"="rebase_per"
              )

              
  )
  plotlyOutput("time")

```

```{r time_server2,context="server"}

output$time <- renderPlotly({
    ghg_yearly<-ghg_allyears %>%
  group_by(Country,Calendar.Year) %>%
    summarise(Annual.CO2.Total=sum(Annual.CO2.Total),
              Mid.year.Population..thousands.=sum(Mid.year.Population..thousands.),
              Area..km2.=sum(Area..km2.))
    
    ghg_yearly<-ghg_yearly %>%

      {if(input$trans_t=="capita")mutate(.,Annual.CO2.Total=(Annual.CO2.Total/Mid.year.Population..thousands.))else . }%>%
      {if(input$trans_t=="area") mutate(.,Annual.CO2.Total=(Annual.CO2.Total/Area..km2.))else . }%>%
    {if(input$trans_t=="rebase") group_by(.,Country) %>%
                        mutate(Annual.CO2.Total=(Annual.CO2.Total-
                                                           mean(Annual.CO2.Total[Calendar.Year==2005],na.rm=T))) else . }%>%
      {if(input$trans_t=="rebase_per") group_by(.,Country) %>%
          mutate(Annual.CO2.Total=-1+(Annual.CO2.Total/
                                             mean(Annual.CO2.Total[Calendar.Year==2005],na.rm=T))) else . }
    
   ( ggplot(ghg_yearly)+
      {if(input$plot_type_time=="line")geom_line(aes(y=Annual.CO2.Total,x=Calendar.Year,colour=Country))}+
      {if(input$plot_type_time=="step")geom_step(aes(y=Annual.CO2.Total,x=Calendar.Year,colour=Country))}+

      {if(input$trans_t=="log") scale_y_log10()}+
    {if(input$trans_t=="sqrt") scale_y_sqrt()}+
       theme_light()+
       scale_x_continuous(breaks=2005:2022)) %>% ggplotly()
    
  })

```

## Summary

In this workbook we have covered a lot of different methods for data
exploration, in not a lot of detail per method. To an extent that
mirrors the whole process of exploratory data analysis - in practice we
want to be quick, and nimble, at producing lots of different summaries
of our data and be able to get some quick insights into whether there
are any problems with our data, how the trends and patterns in our data
look and what directions we may be going towards next.

Depending on your own research questions you may have a clear idea ahead
of time of the theoretical models you want to test from your data; or
you may have a large dataset with many variables and a very open-ended
set of hypotheses. While the amount of time spent on EDA for these two
tasks may be different - the importance of this phase and the sort of
methods that an analyst needs to be aware of is not. And the best way to
get a handle on this phase of the analysis is without doubt to get your
hands dirty by exploring multiple datasets on your own...

## Exercises

Now it's time for you to start exploring some data to see what you can
find. The dataset we will be using concerns earthquakes observed in an
area around Fiji since 1964, obtained from Harvard University Department
of Geophysics. A summary of the variables is below:

```{r}

data.frame(column=c("lat"	,"long","depth","mag","stations","order"),
           details=c("Latitude of earthquake","Longitude of earthquake","Depth (km)",
                     "Richter Scale Magnitude","Number of stations reporting the earthquake",
                     "Order in which the earthquakes occurred"),
           type=c("numeric"))%>%gt::gt()

```

You can access a tool to explore the data here:
<https://shiny.stats4sd.org/earthquakes__dev/>

This will let you view and sort the data, produce summary statistics and
visualisations, and then filter the data or create categories based on
the original variables that could be used in your summary statistics and
visualisations.

If you are trying to improve your skills with any specific data analysis
software or tools (R/Python/Stata etc), then I would strongly encourage
you to download the raw data here: https://raw.githubusercontent.com/stats4sd/explore_earthquakes/refs/heads/dev/quakes.csv and
explore the data using that tool. 

Try to recreate what you have been
able to produce using the interactive app, and then to keep exploring
further beyond those pre-built capabilities!

You are welcome to explore the data in whatever way you see fit;
depending on what you might find interesting or what patterns you start
to uncover.

But to help guide your explorations, here are some questions that a
researcher might have when getting started with a project based on this
data:

-   Does the structure of the data make sense, and does it match what
    would be expected from the description above?
-   Are there certain properties of the data that are unexpected or
    might be problematic?\
-   How could the distribution of the "magnitude" and "depth" variables
    be described?
-   How commonly observed are earthquakes that would be classified as
    “moderate” (5-6 magnitude on the Richter scale) or “strong” (6 and
    above)?
-   Does there look like there relationships between the depth or number
    of stations reporting the earthquake with the magnitude of the
    earthquakes?
-   Are there any trends in the magnitude, the depth, or the number of
    stations reporting each earthquake over time?
-   Are there any patterns in the locations of the earthquakes over
    space, and does the magnitude, depth or number of stations reporting
    vary according to the locations?

## Resources

Tukey, J. W. (1977). Exploratory Data Analysis

A 'classic' text which popularised many of the exploratory techniques
covered in this session, which still makes many relevant general points.

<https://www.stat.cmu.edu/~hseltman/309/Book/chapter4.pdf> A slightly
more technical introduction to EDA then is presented here

<https://r4ds.had.co.nz/exploratory-data-analysis.html>

<https://www.youtube.com/watch?v=QiqZliDXCCg>
