---
title: "Exploratory Data Analysis - Code Only"
author: "Sam Dumble"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}

#install.packages(c("tidyverse","waffle","treemap","Hmisc","psych","agricolae","janitor","gridExtra"))

library(tidyverse)
library(waffle)
library(treemap)
library(Hmisc)
library(psych)
library(agricolae)
library(janitor)
library(gridExtra)
library(plotly)
```

## Data Manipulation

Downloading the file directly from the URL and loading it in

```{r cars}

ghg<-read.csv("https://assets.publishing.service.gov.uk/media/667ad86497ea0c79abfe4bfd/2005-2022-local-authority-ghg-emissions-csv-dataset.csv",stringsAsFactors = TRUE)

```

Two subsets are used in the workbook - one for local authorities in 2022 and one over time.

There are a few tricks in the code here - to be able to calculate the change over time variable, and then to pick out the correct values for total area and total population without double counting.

But there are plenty of other ways of getting to the same point!

```{r}
ghg_2022<-ghg %>%
  #select CO2 and 2022 or 2005
  filter(Greenhouse.gas=="CO2" & Calendar.Year=="2022"|Calendar.Year=="2005") %>%
  #summarise at LA level
  group_by(Country,Country.Code,Region,Region.Code,
           Second.Tier.Authority,Local.Authority,Local.Authority.Code,
           Area..km2.,Mid.year.Population..thousands.,Calendar.Year) %>%
  #get the total across all subsectotrs
  summarise(Annual.CO2.Total=sum(CO2.emissions.within.the.scope.of.influence.of.LAs..kt.CO2.)) %>%
  #remove the unallocated from this analysis
  filter(Country!="Unallocated")%>%
  ungroup() %>%
  #AT LA level determine the % change from 2005
  group_by(Local.Authority)%>%
  mutate(chg=1-(Annual.CO2.Total/mean(Annual.CO2.Total[Calendar.Year=="2005"],na.rm=T)))%>%
  mutate("CO2_Reduction_from_2005"=cut(chg,breaks=c(0,0.3,0.4,0.5,1),labels=c("30% or less",
                                                                                  "30-40%","40-50%","50% or more"))) %>%
  select(-chg) %>%
  #remove the 2005 data
  filter(Calendar.Year=="2022")



ghg_allyears<-ghg %>%
  #first summarise to year by LA level for only the CO2
  filter(Greenhouse.gas=="CO2") %>%
  group_by(Country,Country.Code,Region,Region.Code,
           Second.Tier.Authority,Local.Authority,Local.Authority.Code,
           Calendar.Year,Area..km2.,Mid.year.Population..thousands.) %>%
  summarise(Annual.CO2.Total=sum(CO2.emissions.within.the.scope.of.influence.of.LAs..kt.CO2.)) %>%
  filter(Country!="Unallocated")%>%
  #Now that we have one row per LA per year we can sum the area and population within each country
  #If we did this in previous step we would be counting same values multiple times
  group_by(Country,Calendar.Year) %>%
    summarise(Annual.CO2.Total=sum(Annual.CO2.Total),
              Mid.year.Population..thousands.=sum(Mid.year.Population..thousands.),
              Area..km2.=sum(Area..km2.))

```

### Looking at the UK Greenhouse Gas Data


```{r}
ghg %>%
  slice(1:1000)
```

```{r}
summary(ghg)
```

```{r}
ghg %>%
  group_by(Country) %>%
    summarise(Total=n(),"Population Recorded"=sum(!is.na(Mid.year.Population..thousands.)),
              "Population Missing"=sum(is.na(Mid.year.Population..thousands.)),
              "Area Recorded"=sum(!is.na(Area..km2.)),
              "Area Missing"=sum(is.na(Area..km2.)))
```

### Summary Statistics

```{r}
ghg_2022 %>%
  ungroup() %>%
      summarise("Number of Counties"=n(),
                Mean=mean(Annual.CO2.Total)%>%round(1),
                Median=median(Annual.CO2.Total)%>%round(1),
                "Trimmed mean"=mean(Annual.CO2.Total,trim = 0.05)%>%round(1),
                 "Geometric mean"=geometric.mean(Annual.CO2.Total)%>%round(1),
                "Harmonic mean"=harmonic.mean(Annual.CO2.Total)%>%round(1),
                 Sum=sum(Annual.CO2.Total)%>%round(),
                "Minimum"=min(Annual.CO2.Total)%>%round(1),
                "Maximum"=max(Annual.CO2.Total)%>%round(1),
                "Range"=(max(Annual.CO2.Total)-min(Annual.CO2.Total))%>%round(1),
                "Standard Deviation"=sd(Annual.CO2.Total)%>%round(1),
                "Variance"=var(Annual.CO2.Total)%>%round(),
                "Coefficient of Variation"=scales::percent(sd(Annual.CO2.Total)/mean(Annual.CO2.Total)),
                "Lower Quartile"=quantile(Annual.CO2.Total,0.25)%>%round(1),
                "Upper Quartile"=quantile(Annual.CO2.Total,0.75)%>%round(1),
                "Inter Quartile Range"=diff(quantile(Annual.CO2.Total,c(0.25,0.75)))%>%round(1),
                "Median Absolute Deviation"=mad(Annual.CO2.Total)%>%round(1),
                "Kurtosis"=kurtosis(Annual.CO2.Total)%>%round(2),
                "Skewness"=skewness(Annual.CO2.Total)%>%round(2),
                "% of Values Over Threshold (100kt)"=scales::percent(mean(Annual.CO2.Total>1000)),
                )%>%
  gather(Statistic,Value)
```
## One variable plots

'One variable' but often here I have cheated by putting in two variables as ggplot will require two variables to plot. As Calendar Year is constant (always 2022) then this is not really a variable!

Minimal examples of the code included without any formatting: in EDA our primary goal is to make things quickly/flexiblly & functionally. Making it look pretty can wait until later!

```{r}
 ggplot(ghg_2022)+
   geom_point(aes(y=Annual.CO2.Total,x=Calendar.Year))
```

```{r}
 ggplot(ghg_2022)+
  geom_point(aes(y=Annual.CO2.Total,x=rank(Annual.CO2.Total)))
```

```{r}
 ggplot(ghg_2022)+
     geom_jitter(aes(y=Annual.CO2.Total,x=Calendar.Year),height=0,width=0.25)
```

```{r}
 ggplot(ghg_2022)+
 stat_summary(aes(y=Annual.CO2.Total,x=Calendar.Year),fun.data=mean_cl_normal,conf.int=0.95)
```

```{r}
 ggplot(ghg_2022)+
geom_histogram(aes(x=Annual.CO2.Total))
```


```{r}
 ggplot(ghg_2022)+
geom_freqpoly(aes(x=Annual.CO2.Total))
```

```{r}
 ggplot(ghg_2022)+
geom_violin(aes(y=Annual.CO2.Total,x=Calendar.Year,group=Calendar.Year))
```

```{r}
 ggplot(ghg_2022)+
geom_density(aes(x=Annual.CO2.Total))
```

```{r}
 ggplot(ghg_2022)+
geom_boxplot(aes(y=Annual.CO2.Total,x=Calendar.Year,group=Calendar.Year))
```

```{r}
 ggplot(ghg_2022)+
  stat_summary(geom="col",fun="mean",aes(y=Annual.CO2.Total,x=Calendar.Year))
```


```{r}
 ggplot(ghg_2022)+
stat_ecdf(geom = "step",aes(x=Annual.CO2.Total))
```


```{r}
 ggplot(ghg_2022)+
  stat_qq(aes(sample=Annual.CO2.Total))+
  stat_qq_line(aes(sample=Annual.CO2.Total))
```

## Categorical variables

```{r}
ghg_2022 %>%
  group_by(CO2_Reduction_from_2005) %>%
      summarise("Number of Counties"=n(),
                "Percentage of Counties"=scales::percent(n()/nrow(ghg_2022),0.1))
```

Producing summary table first and then using this to create plots. To plot percentages/proportions then simply swap out "n" for "p"

```{r}
 sum_tab<-   ghg_2022 %>%
  group_by(CO2_Reduction_from_2005) %>%
      summarise(n=n(),
                p=n()/nrow(ghg_2022))
```

```{r}
ggplot(sum_tab,aes(x=CO2_Reduction_from_2005,y=n,fill=CO2_Reduction_from_2005))+
          geom_col()
```

```{r}
ggplot(ghg_2022,aes(y=CO2_Reduction_from_2005,x=1,col=CO2_Reduction_from_2005))+
          geom_count()
```

```{r}
ggplot(sum_tab,aes(x=1,y=n,fill=CO2_Reduction_from_2005))
          geom_col()
```

```{r}
waffle(select(sum_tab,CO2_Reduction_from_2005,n),
       rows=19, size=0.5,title = "CO2 Reduction Across Counties Within The UK",
       xlab="1 Square  = 1 County") 
```

```{r}
ggplot(sum_tab,aes(x=1,fill=CO2_Reduction_from_2005,y=n))+
  geom_col()+
  coord_polar("y", start=0)

```

```{r}

treemap(sum_tab,index="CO2_Reduction_from_2005",vSize="n")  
```

Two variables

```{r}
ghg_2022 %>%
  group_by(Country) %>%
      summarise("Number of Counties"=n(),
                Mean=mean(Annual.CO2.Total)%>%round(1),
                Median=median(Annual.CO2.Total)%>%round(1),
                "Trimmed mean"=mean(Annual.CO2.Total,trim = 0.05)%>%round(1),
                 "Geometric mean"=psych::geometric.mean(Annual.CO2.Total)%>%round(1),
                "Harmonic mean"=psych::harmonic.mean(Annual.CO2.Total)%>%round(1),
                 Sum=sum(Annual.CO2.Total)%>%round(),
                "Minimum"=min(Annual.CO2.Total)%>%round(1),
                "Maximum"=max(Annual.CO2.Total)%>%round(1),
                "Range"=(max(Annual.CO2.Total)-min(Annual.CO2.Total))%>%round(1),
                "Standard Deviation"=sd(Annual.CO2.Total)%>%round(1),
                "Variance"=var(Annual.CO2.Total)%>%round(),
                "Coefficient of Variation"=scales::percent(sd(Annual.CO2.Total)/mean(Annual.CO2.Total)),
                "Lower Quartile"=quantile(Annual.CO2.Total,0.25)%>%round(1),
                "Upper Quartile"=quantile(Annual.CO2.Total,0.75)%>%round(1),
                "Inter Quartile Range"=diff(quantile(Annual.CO2.Total,c(0.25,0.75)))%>%round(1),
                "Median Absolute Deviation"=mad(Annual.CO2.Total)%>%round(1),
                "Kurtosis"=agricolae::kurtosis(Annual.CO2.Total)%>%round(2),
                "Skewness"=agricolae::skewness(Annual.CO2.Total)%>%round(2),
                "% of Values Over Threshold (100kt)"=scales::percent(mean(Annual.CO2.Total>1000)),
                )%>%
  gather(Statistic,Value,-Country ,factor_key = TRUE) %>%
  spread(Country ,Value)
```

## Plots with two variables

Points:
```{r}
 ggplot(ghg_2022)+
   geom_point(aes(y=Annual.CO2.Total,x=Country))
```

Caterpillars:
```{r}
 ggplot(ghg_2022)+
  geom_point(aes(y=Annual.CO2.Total,x=rank(Annual.CO2.Total)))+
  facet_wrap(~Country)
```
Jittered points:
```{r}
 ggplot(ghg_2022)+
     geom_jitter(aes(y=Annual.CO2.Total,x=Country),height=0,width=0.25)
```
Error bars:
```{r}
 ggplot(ghg_2022)+
 stat_summary(aes(y=Annual.CO2.Total,x=Country),fun.data=mean_cl_normal,conf.int=0.95)
```

histograms:
```{r}
 ggplot(ghg_2022)+
geom_histogram(aes(x=Annual.CO2.Total))+
  facet_wrap(~Country)
```

Frequecncy Polygon:
```{r}
 ggplot(ghg_2022)+
geom_freqpoly(aes(x=Annual.CO2.Total,colour=Country))
```
Violins:
```{r}
 ggplot(ghg_2022)+
geom_violin(aes(y=Annual.CO2.Total,x=Country))
```

Density plots:
```{r}
 ggplot(ghg_2022)+
geom_density(aes(x=Annual.CO2.Total,color=Country))
```

Boxplots:
```{r}
 ggplot(ghg_2022)+
geom_boxplot(aes(y=Annual.CO2.Total,x=Country))
```
Bar charts:
```{r}
 ggplot(ghg_2022)+
  stat_summary(geom="col",fun="mean",aes(y=Annual.CO2.Total,x=Country))
```

Cumulative density:
```{r}
 ggplot(ghg_2022)+
stat_ecdf(geom = "step",aes(x=Annual.CO2.Total,color=Country))
```

QQ Plots:
```{r}
 ggplot(ghg_2022)+
  stat_qq(aes(sample=Annual.CO2.Total))+
  stat_qq_line(aes(sample=Annual.CO2.Total))+
  facet_wrap(~Country)
```

### Cross Tabs

The `janitor` package has a nice interface for switching between the different percentage types in a more intuitive way than using dplyr.

```{r}
ghg_2022 %>%
  tabyl(CO2_Reduction_from_2005,Country) %>%
      adorn_percentages(denominator = "row")%>%
      adorn_pct_formatting() %>%
      adorn_ns(format_func = function(x)paste("n =",x))

ghg_2022 %>%
  tabyl(CO2_Reduction_from_2005,Country) %>%
      adorn_percentages(denominator = "col")%>%
      adorn_pct_formatting() %>%
      adorn_ns(format_func = function(x)paste("n =",x))

ghg_2022 %>%
  tabyl(CO2_Reduction_from_2005,Country) %>%
      adorn_percentages(denominator = "all")%>%
      adorn_pct_formatting() %>%
      adorn_ns(format_func = function(x)paste("n =",x))
```

```{r}
sum_tab2<-   ghg_2022 %>%
  group_by(Country,CO2_Reduction_from_2005) %>%
      summarise(n=n()) %>%
  mutate(p=n/sum(n))%>%ungroup()%>%
  rbind(data.frame(Country="Wales",
                   CO2_Reduction_from_2005="30% or less",n=0,p=0))%>%
  rbind(data.frame(Country="Northern Ireland",  CO2_Reduction_from_2005="50% or more",n=0,p=0))%>%
  arrange(Country,CO2_Reduction_from_2005)

sum_tab2$CO2_Reduction_from_2005<-factor(sum_tab2$CO2_Reduction_from_2005)
```

Bars
```{r}
ggplot(sum_tab2,aes(x=Country,y=p,fill=CO2_Reduction_from_2005))+
          geom_col(show.legend = FALSE,col="black",position="dodge")
```

Stacked Bars
```{r}

ggplot(sum_tab2,aes(x=Country,y=p,fill=CO2_Reduction_from_2005))+
          geom_col(show.legend = TRUE,col="black")
```

Pie chart
```{r}
ggplot(sum_tab2,aes(x=1,fill=CO2_Reduction_from_2005,y=p))+
  geom_col(col="black",alpha=0.8)+
  coord_polar("y", start=0)+
   facet_wrap(~Country,scales="free")
```

Treemaps
```{r}
treemap(sum_tab2,
            index=c("Country","CO2_Reduction_from_2005"),
            vSize="p",
            type="index",
            vColor="CO2_Reduction_from_2005")
            
```

Waffle plots: (not really a way to do this with multiple variables easily, so having to take apart and then put back together!)
```{r}

p1.1<-waffle(select(filter(sum_tab2,Country=="England"),CO2_Reduction_from_2005,p)%>%
                mutate(p=round(100*p)),
       rows=10, size=0.5,title = "England",keep = TRUE,
       xlab="1 Square  = 1 %")
p1.2<-waffle(select(filter(sum_tab2,Country=="Scotland"),CO2_Reduction_from_2005,p)%>%
                mutate(p=round(100*p)),
       rows=10, size=0.5,title = "Scotland",keep = TRUE,
       xlab="1 Square  = 1 %")
p1.3<-waffle(select(filter(sum_tab2,Country=="Wales"),CO2_Reduction_from_2005,p)%>%
                mutate(p=round(100*p)),
       rows=10, size=0.5,title = "Wales",keep=TRUE ,
       xlab="1 Square  = 1 %")
p1.4<-waffle(select(filter(sum_tab2,Country=="Northern Ireland"),CO2_Reduction_from_2005,p)%>%
                mutate(p=round(100*p)),
       rows=10, size=0.5,title = "Northern Ireland",keep = TRUE,
       xlab="1 Square  = 1 %")


 marrangeGrob(list(p1.1,p1.2,p1.3,p1.4), nrow=2, ncol=2,top = "Percentages") %>% print()
      

```

Count plot

```{r}
ggplot(ghg_2022,aes(y=CO2_Reduction_from_2005,x=Country,group=Country))+
          geom_count(show.legend = TRUE,col="black",shape=21,aes(size = after_stat(prop),fill=CO2_Reduction_from_2005))
```


## Two Numeric Variables

```{r}
ggplot(data=ghg_2022,aes(x=Mid.year.Population..thousands.,y=Annual.CO2.Total))+
  geom_point()
```

```{r}
ggplot(data=ghg_2022,aes(x=Mid.year.Population..thousands.,y=Annual.CO2.Total))+
  geom_point()+
  geom_smooth(method="lm")

ggplot(data=ghg_2022,aes(x=Mid.year.Population..thousands.,y=Annual.CO2.Total))+
  geom_point()+
  geom_smooth(method="loess")

ggplot(data=ghg_2022,aes(x=Mid.year.Population..thousands.,y=Annual.CO2.Total))+
  geom_point()+
  geom_smooth(method="gam")
```


As an extra 'type' of plot - the ability to identify points through interactivity is very useful!

Can make any ggplot interactive using ggplotly, and adding any variables to the aes() will enable them when you click/hover over the points

```{r}
p1<-ggplot(data=ghg_2022,aes(x=Mid.year.Population..thousands.,y=Annual.CO2.Total,Local.Authority=Local.Authority))+
  geom_point()
ggplotly(p1)
```




```{r}
ghg_2022 %>%
  ungroup() %>%
  select(Area..km2.,Mid.year.Population..thousands.,Annual.CO2.Total) %>%
      cor(method="pearson")

ghg_2022 %>%
  ungroup() %>%
  select(Area..km2.,Mid.year.Population..thousands.,Annual.CO2.Total) %>%
      cor(method="spearman")

ghg_2022 %>%
  ungroup() %>%
  select(Area..km2.,Mid.year.Population..thousands.,Annual.CO2.Total) %>%
      cor(method="kendall")
```

I am not going to include every possible scatter plot you could make from the tool! But will be some combination of variables like this:
```{r}
ggplot(ghg_2022,aes(y=Annual.CO2.Total,x=Area..km2.))+
    geom_point(aes(colour=Country,size=Mid.year.Population..thousands.),alpha=0.6)+
scale_x_log10()+
 scale_y_log10()
```



#Trends over time

Calculating the transformed and derived variables

```{r}
  ghg_allyears<-ghg_allyears %>%
  mutate(Annual.CO2.Total_percap=(Annual.CO2.Total/Mid.year.Population..thousands.))%>%
mutate(Annual.CO2.Total_perkm=(Annual.CO2.Total/Area..km2.))%>%
group_by(Country) %>%
  mutate(Annual.CO2.Total_rebase=(Annual.CO2.Total-mean(Annual.CO2.Total[Calendar.Year==2005],na.rm=T)))%>%
mutate(Annual.CO2.Total_rebase_per=-1+(Annual.CO2.Total/
                                             mean(Annual.CO2.Total[Calendar.Year==2005],na.rm=T)))

```

Then making plots with either step or line:

```{r}
    
ggplot(ghg_allyears)+geom_line(aes(y=Annual.CO2.Total_percap,x=Calendar.Year,colour=Country))
```

```{r}

ggplot(ghg_allyears)+geom_step(aes(y=Annual.CO2.Total_percap,x=Calendar.Year,colour=Country))
```

